---
title: "Agrobusiness"
author: "Kate, Anastasiia, Max"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 7, dpi = 150)
```

------------------------------------------------------------------------

# Hypotheses and objectives description.

Our project researches the "story" of wheat market volatility in Ukraine. We are solving the problem of agricultural risk management by identifying which meteorological and human-led factors (like fertilizers) actually drive yields and prices. We aim to provide agribusinesses with a clear framework for data collection: what range of data is mandatory and what is redundant.

### Research Hypotheses

**1. Traditional Significance and Input Factors:**

$H_0$: Meteorological factors (radiation, temperature, humidity), fertilizers, and chemicals have no statistically significant effect on wheat yield or prices.

$H_1$: These factors have a significant impact, specifically following a non-linear "U-shaped" or parabolic curve where both extremes are detrimental.

**2. Seasonal Relevancy (The "Winter" Question):**

$H_0$: Weather data from outside the active spring/summer growing season (e.g., winter temperatures) is not significant for predicting harvest outcomes.

$H_1$: Winter weather is a significant predictor of soil moisture reserves and winter-kill rates, which directly impact the final yield and market price dynamics.

**3. Data Granularity and Extreme Day Classification:** - **Goal:** We test a custom classification where average daily temperature $(T_{max} - T_{min}) / 2$ is used to categorize days as "Extremely Cold" ($< -10^{\circ}C$), "Cold" ($-10^{\circ}C$ to $5^{\circ}C$), "Warm" ($5^{\circ}C$ to $30^{\circ}C$), or "Hot" ($> 30^{\circ}C$).

$H_0$: Summing "extra days" (extreme heat/cold) does not provide more predictive power than simple weekly averages.

$H_1$: The frequency of extreme temperature events is a more accurate predictor of yield shocks than mean temperature alone, requiring high-frequency (daily) data collection for valid modeling.

**4. Cloudiness and Correlation Logic (Variable Selection):**

$H_0$: Cloudiness, rain, and sunshine duration are independent predictors with no significant correlation.

$H_1$: Cloudiness is strongly correlated with precipitation and solar radiation. We hypothesize that using all three creates multicollinearity; therefore, we test which single variable (e.g., "shine") serves as the most effective proxy for the model.

------------------------------------------------------------------------

# Data analysis.

Provide your data analysis. What conclusions can be drawn from this analysis? Does this analysis lead to any first conclusions about the research question or suggest any hypotheses worth testing?

**Datasets:**

-   sowing_area.csv

-   harvested_area.csv

-   crop_capacity.csv

-   daily_weather.csv

-   soil_data.csv

------------------------------------------------------------------------

# Data Description and Exploratory Analysis

## 1. Introduction

This study investigates the impact of weather factors on winter wheat yield in Ukraine using panel data for 12 years (2013–2024) at the oblast level. The dependent variable is yield (centners per hectare). Key explanatory variables are derived from daily temperature data following a theoretically motivated aggregation scheme. Soil properties and area variables serve as controls.

------------------------------------------------------------------------

## 2. Data Description

This section describes all data sources used in the study: crop statistics (annual, oblast level), daily weather records, static soil properties, and derived weather metrics constructed following the research hypothesis.

### 2.1 Crop Data (annual, by oblast)

| File | Variable | Units | Explanation |
|------------------|------------------|------------------|------------------|
| `crop_capacity.csv` | Yield | centners/ha | Dependent variable. Already calculated by statistical authorities. |
| `harvested_area.csv` | Harvested area | hectares | Area from which crop was actually harvested. |
| `sowing_area.csv` | Sown area | hectares | Area planted (may differ from harvested due to losses). |

**Notes:** One centner = 100 kg. Typical winter wheat yield in Ukraine ranges from 30–80 ц/га. The difference between sown and harvested area indicates crop losses during vegetation.

```{r}
# ============================================================
# Interim Report – Part 1
# Winter Wheat Yield in Ukraine: Data Description & EDA
# Updated for new file format (2015–2025, thousands of ha)
# ============================================================

# ── 0. Libraries ─────────────────────────────────────────────
library(tidyverse)
library(lubridate)
library(patchwork)

# ── 1. Load data ─────────────────────────────────────────────
data_path <- "."

# New files have 8 leading metadata columns before the year columns:
# "Показник", "Територіальний розріз", "Одиниця виміру",
# "Масштабування", "Кількість десяткових знаків",
# "Спеціальне значення даних", "Примітки часового ряду",
# "Примітки спостереження", then 2015..2025

years <- as.character(2015:2025)

META_COLS <- c("indicator", "region", "unit", "scale",
               "decimals", "special", "note_ts", "note_obs")

read_panel <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names = c(META_COLS, years),
    skip       = 1,            # skip original header row
    show_col_types = FALSE
  ) |>
    # keep only region name and year columns
    select(region, all_of(years)) |>
    pivot_longer(
      cols      = all_of(years),
      names_to  = "year",
      values_to = value_name
    ) |>
    mutate(
      year         = as.integer(year),
      # replace empty strings and obvious placeholders with NA
      !!value_name := suppressWarnings(as.numeric(.data[[value_name]]))
    ) |>
    # drop the "Україна" aggregate and years with no data at all
    filter(region != "Україна")
}

# Yield: centners / ha  (dependent variable — already computed)
yield <- read_panel("harvest_capacity.csv", "yield_cpha")

# Harvested area: thousands of ha  → convert to ha for consistency
harv  <- read_panel("harvested_area.csv",  "harv_kha") |>
  mutate(harv_ha = harv_kha * 1000) |>
  select(-harv_kha)

# Sown area: thousands of ha  → convert to ha
sown  <- read_panel("sowing_area.csv",     "sown_kha") |>
  mutate(sown_ha = sown_kha * 1000) |>
  select(-sown_kha)

# Keep Ukraine aggregate separately for reference lines
read_ukraine <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names = c(META_COLS, years),
    skip       = 1,
    show_col_types = FALSE
  ) |>
    filter(.data[["region"]] == "Україна") |>
    select(region, all_of(years)) |>
    pivot_longer(cols = all_of(years), names_to = "year",
                 values_to = value_name) |>
    mutate(year = as.integer(year),
           !!value_name := suppressWarnings(as.numeric(.data[[value_name]])))
}

ukr_yield <- read_ukraine("harvest_capacity.csv", "yield_cpha")
ukr_harv  <- read_ukraine("harvested_area.csv",  "harv_kha")
ukr_sown  <- read_ukraine("sowing_area.csv",     "sown_kha")

# ── 2. Quick data check ───────────────────────────────────────
cat("Yield: ", nrow(yield), "rows |",
    n_distinct(yield$region), "regions |",
    paste(range(yield$year, na.rm = TRUE), collapse = "–"), "\n")
cat("Sample yield values (Vinnytsia):\n")
print(yield |> filter(region == "Вінницька") |> select(year, yield_cpha))

# ── 3. Region groups ──────────────────────────────────────────
region_groups <- tribble(
  ~region,              ~group,
  # West
  "Волинська",          "West",
  "Закарпатська",       "West",
  "Івано-Франківська",  "West",
  "Львівська",          "West",
  "Рівненська",         "West",
  "Тернопільська",      "West",
  "Хмельницька",        "West",
  "Чернівецька",        "West",
  # Centre
  "Вінницька",          "Centre",
  "Дніпропетровська",   "Centre",
  "Житомирська",        "Centre",
  "Київська",           "Centre",
  "Кіровоградська",     "Centre",
  "Полтавська",         "Centre",
  "Сумська",            "Centre",
  "Харківська",         "Centre",
  "Черкаська",          "Centre",
  "Чернігівська",       "Centre",
  # East / South
  "Донецька",           "East",
  "Запорізька",         "East",
  "Луганська",          "East",
  "Миколаївська",       "East",
  "Одеська",            "East",
  "Херсонська",         "East"
)

# ── 4. Colour palette ─────────────────────────────────────────
GROUP_COLORS <- c(West = "#2c7bb6", Centre = "#d7191c", East = "#1a9641")

# ── 5. Shared theme ───────────────────────────────────────────
theme_small_multiple <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major   = element_line(colour = "grey92", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      panel.border       = element_blank(),
      panel.background   = element_rect(fill = "white", colour = NA),
      plot.background    = element_rect(fill = "white", colour = NA),
      axis.line.x        = element_line(colour = "grey70", linewidth = 0.4),
      axis.ticks.x       = element_line(colour = "grey70", linewidth = 0.3),
      axis.ticks.y       = element_blank(),
      axis.text.x        = element_text(size = 7, colour = "grey40",
                                        margin = margin(t = 2)),
      axis.text.y        = element_text(size = 7, colour = "grey40"),
      axis.title         = element_blank(),
      strip.text         = element_text(face = "bold", size = 8,
                                        colour = "grey20",
                                        margin = margin(b = 3)),
      strip.background   = element_blank(),
      strip.placement    = "outside",
      legend.position    = "none",
      plot.title         = element_text(face = "bold", size = 11,
                                        colour = "grey10",
                                        margin = margin(b = 4)),
      plot.subtitle      = element_text(size = 8.5, colour = "grey40",
                                        margin = margin(b = 8)),
      plot.caption       = element_text(size = 7, colour = "grey55",
                                        hjust = 0, margin = margin(t = 6)),
      plot.margin        = margin(10, 12, 8, 10)
    )
}

# ── 6. Small-multiple helper ──────────────────────────────────
# Each facet = one focal region (highlighted in group colour).
# All other regions shown as thin grey background lines.

plot_small_multiple <- function(df, group_map, value_col,
                                group    = NULL,
                                title    = "",
                                subtitle = "",
                                caption  = "") {

  df <- df |>
    left_join(group_map, by = "region") |>
    rename(value = !!sym(value_col))

  focal_regions <- if (!is.null(group)) {
    group_map |> filter(group == !!group) |> pull(region)
  } else {
    unique(df$region)
  }

  focal_color <- if (!is.null(group)) GROUP_COLORS[[group]] else "grey30"

  # Background: every other region repeated across all focal facets
  bg_data <- tibble(facet = focal_regions) |>
    cross_join(
      df |> filter(!is.na(value)) |> select(region, year, value)
    )

  # Focal: one line per facet
  focal_data <- df |>
    filter(region %in% focal_regions, !is.na(value)) |>
    mutate(facet = region)

  # Last valid point per facet for endpoint dot
  endpoint <- focal_data |>
    group_by(facet) |>
    slice_max(year, n = 1, with_ties = FALSE) |>
    ungroup()

  ggplot() +
    geom_line(
      data    = bg_data |> filter(region != facet),
      mapping = aes(x = year, y = value, group = region),
      colour  = "grey85", linewidth = 0.35, alpha = 0.9
    ) +
    geom_line(
      data    = focal_data,
      mapping = aes(x = year, y = value, group = facet),
      colour  = focal_color, linewidth = 1.1, lineend = "round"
    ) +
    geom_point(
      data    = endpoint,
      mapping = aes(x = year, y = value),
      colour  = focal_color, size = 2.2,
      shape = 21, fill = "white", stroke = 1.4
    ) +
    facet_wrap(~facet, scales = "free_y") +
    scale_x_continuous(
      breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
      labels = c("'15", "'17", "'19", "'21", "'23", "'25")
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
    labs(title = title, subtitle = subtitle, caption = caption) +
    theme_small_multiple()
}

# ── 7. Yield plots ────────────────────────────────────────────
p_yield_west <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "West",
  title    = "Winter wheat yield — Western oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_yield_centre <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "Centre",
  title    = "Winter wheat yield — Central oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_yield_east <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "East",
  title    = "Winter wheat yield — Eastern oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── 8. Harvested area plots ───────────────────────────────────
p_harv_west <- plot_small_multiple(
  harv, region_groups, "harv_ha", "West",
  title    = "Harvested area — Western oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_harv_centre <- plot_small_multiple(
  harv, region_groups, "harv_ha", "Centre",
  title    = "Harvested area — Central oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_harv_east <- plot_small_multiple(
  harv, region_groups, "harv_ha", "East",
  title    = "Harvested area — Eastern oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── 9. Sown area plots ────────────────────────────────────────
p_sown_west <- plot_small_multiple(
  sown, region_groups, "sown_ha", "West",
  title    = "Sown area — Western oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_sown_centre <- plot_small_multiple(
  sown, region_groups, "sown_ha", "Centre",
  title    = "Sown area — Central oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_sown_east <- plot_small_multiple(
  sown, region_groups, "sown_ha", "East",
  title    = "Sown area — Eastern oblasts",
  subtitle = "Hectares, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── 10. Representative oblasts comparison ─────────────────────
# One oblast per group with highest mean yield → overlay all three
# against the Ukraine national average

rep_oblasts <- yield |>
  left_join(region_groups, by = "region") |>
  filter(!is.na(group), !is.na(yield_cpha)) |>
  group_by(group, region) |>
  summarise(mean_y = mean(yield_cpha, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_y, n = 1, by = group)

rep_data <- yield |>
  inner_join(rep_oblasts |> select(group, region), by = "region") |>
  filter(!is.na(yield_cpha))

p_representative <- ggplot() +
  # Ukraine national average (dashed grey)
  geom_line(
    data    = ukr_yield |> filter(!is.na(yield_cpha)),
    mapping = aes(x = year, y = yield_cpha),
    colour  = "grey65", linewidth = 0.7, linetype = "dashed"
  ) +
  # Representative oblast line
  geom_line(
    data    = rep_data,
    mapping = aes(x = year, y = yield_cpha, colour = group),
    linewidth = 1.2, lineend = "round"
  ) +
  # Endpoint dots
  geom_point(
    data = rep_data |>
      group_by(group) |>
      slice_max(year, n = 1, with_ties = FALSE) |>
      ungroup(),
    mapping = aes(x = year, y = yield_cpha, colour = group),
    size = 2.5, shape = 19
  ) +
  # Oblast name labels on right side
  geom_text(
    data = rep_oblasts,
    mapping = aes(label = paste0(region, "\n(", group, ")")),
    x = Inf, y = Inf,
    hjust = 1.05, vjust = 1.3,
    size = 2.8, colour = "grey30"
  ) +
  facet_wrap(~group) +
  scale_colour_manual(values = GROUP_COLORS) +
  scale_x_continuous(
    breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
    labels = c("'15", "'17", "'19", "'21", "'23", "'25")
  ) +
  labs(
    title    = "Most productive representative oblast per geographical group",
    subtitle = "Dashed grey = Ukraine national average. Coloured = representative oblast (highest mean yield).",
    caption  = "Source: State Statistics Service of Ukraine"
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold", size = 10, colour = "grey15")
  )

# ── 11. Print all ─────────────────────────────────────────────
print(p_yield_west)
print(p_yield_centre)
print(p_yield_east)
print(p_harv_west)
print(p_harv_centre)
print(p_harv_east)
print(p_sown_west)
print(p_sown_centre)
print(p_sown_east)
print(p_representative)

# ── 12. Save (uncomment as needed) ────────────────────────────
# ggsave("yield_west.png",    p_yield_west,    width=12, height=8, dpi=300, bg="white")
# ggsave("yield_centre.png",  p_yield_centre,  width=12, height=8, dpi=300, bg="white")
# ggsave("yield_east.png",    p_yield_east,    width=12, height=8, dpi=300, bg="white")
# ggsave("harv_west.png",     p_harv_west,     width=12, height=8, dpi=300, bg="white")
# ggsave("harv_centre.png",   p_harv_centre,   width=12, height=8, dpi=300, bg="white")
# ggsave("harv_east.png",     p_harv_east,     width=12, height=8, dpi=300, bg="white")
# ggsave("sown_west.png",     p_sown_west,     width=12, height=8, dpi=300, bg="white")
# ggsave("sown_centre.png",   p_sown_centre,   width=12, height=8, dpi=300, bg="white")
# ggsave("sown_east.png",     p_sown_east,     width=12, height=8, dpi=300, bg="white")
# ggsave("representative.png",p_representative,width=12, height=5, dpi=300, bg="white")

```

### 2.2 Daily Weather Data (`daily_weather.csv`)

| Variable | Description | Unit |
|------------------------|------------------------|------------------------|
| `region` | Oblast name | text |
| `date` | Calendar date | YYYY-MM-DD |
| `temperature_2m_max` | Daily maximum temperature | °C |
| `temperature_2m_min` | Daily minimum temperature | °C |
| `temperature_2m_mean` | Daily mean temperature | °C |
| `wind_speed_10m_max` | Daily maximum wind speed at 10m height | m/s |
| `wind_gusts_10m_max` | Daily maximum wind gusts at 10m height | m/s |
| `et0_fao_evapotranspiration` | Reference evapotranspiration (FAO Penman-Monteith) | mm/day |
| `dewpoint_2m_mean` | Daily mean dewpoint temperature | °C |
| `precipitation_sum` | Daily total precipitation | mm |
| `shortwave_radiation_sum` | Daily incoming shortwave solar radiation | MJ/m² |

**Notes:** ET0 represents atmospheric evaporative demand — higher values indicate faster water loss from crops. Dewpoint is a direct measure of air moisture (higher values promote fungal diseases). Data cover 2013–2024 for all 24 oblasts.

#### Constructed Weather Metrics (following the research hypothesis)

For each region and year, daily temperature is transformed as:

$$ t_{dev} = \frac{T_{max} + T_{min}}{2} - 5 $$

where 5°C is the assumed base temperature for winter wheat development.

From this deviation, four metrics are calculated for the **full year** (later the vegetation period will be tested):

| Metric | Definition |
|------------------------------------|------------------------------------|
| `days_hot_critical` | Number of days with $t_{dev} > 20$ (extreme heat stress) |
| `days_cold_critical` | Number of days with $t_{dev} < -15$ (extreme cold stress) |
| `sum_heat_noncrit` | Sum of $t_{dev}$ for days with $0 < t_{dev} \leq 20$ |
| `sum_cold_noncrit` | Sum of $t_{dev}$ for days with $-15 \leq t_{dev} < 0$ |

**Notes:** These variables capture different aspects of temperature stress: critical extremes (counts) and accumulated thermal resources (sums). Thresholds (20°C, -15°C) are based on agronomic literature for winter wheat.

```{r fig.width=14, fig.height=9}
# ── Load and aggregate ─────────────────────────────────────
weather <- read_csv("daily_weather.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date), year = year(date)) |>
  filter(year >= 2013, year <= 2024)

BASE_TEMP <- 5

weather_annual <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - BASE_TEMP,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20,  t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0,  t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    days_hot_crit    = sum(days_hot_crit,  na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit, na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,   na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,   na.rm = TRUE),
    mean_tmax        = mean(temperature_2m_max,          na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min,          na.rm = TRUE),
    mean_temp        = mean(t_mean,                      na.rm = TRUE),
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    mean_et0         = mean(et0_fao_evapotranspiration,  na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    .groups          = "drop"
  )

# Order regions W→E by mean temperature (warmer = further east/south)
region_order <- weather_annual |>
  group_by(region) |>
  summarise(mt = mean(mean_temp)) |>
  arrange(mt) |>
  pull(region)

weather_annual <- weather_annual |>
  mutate(region = factor(region, levels = region_order),
         year   = factor(year))
# ── Shared theme ────────────────────────────────────────────
theme_heatmap <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid        = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.text.x       = element_text(size = 7.5, angle = 45, hjust = 1,
                                       colour = "grey35"),
      axis.text.y       = element_text(size = 7,   colour = "grey30"),
      axis.title        = element_blank(),
      axis.ticks        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8.5,
                                       colour = "grey15"),
      strip.background  = element_blank(),
      plot.title        = element_text(face = "bold", size = 10,
                                       colour = "grey10",
                                       margin = margin(b = 2)),
      plot.subtitle     = element_text(size = 7.5, colour = "grey45",
                                       margin = margin(b = 5)),
      legend.text       = element_text(size = 7,   colour = "grey35"),
      legend.title      = element_blank(),
      legend.key.width  = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin       = margin(6, 8, 6, 6)
    )
}

# ── Helper: one heatmap tile ────────────────────────────────
make_tile <- function(df, metric, palette, direction = 1,
                      title, subtitle) {
  ggplot(df, aes(x = year, y = region, fill = .data[[metric]])) +
    geom_tile(colour = "white", linewidth = 0.35) +
    scale_fill_distiller(palette   = palette,
                         direction = direction,
                         guide     = guide_colorbar(
                           barwidth  = 0.45,
                           barheight = 7,
                           ticks     = FALSE)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(title = title, subtitle = subtitle) +
    theme_heatmap()
}
p1 <- make_tile(weather_annual, "days_hot_crit", "YlOrRd", 1,
  "Days with extreme heat stress",
  "t_dev > 20 °C  (mean daily temp > 25 °C) — count of days per year")

p2 <- make_tile(weather_annual, "days_cold_crit", "Blues", -1,
  "Days with extreme cold stress",
  "t_dev < −15 °C  (mean daily temp < −10 °C) — count of days per year")

p3 <- make_tile(weather_annual, "sum_heat_noncrit", "YlOrBr", 1,
  "Accumulated non-critical heat  (growing degree-days)",
  "Sum of t_dev for days where 0 < t_dev ≤ 20 °C — thermal budget for growth")

p4 <- make_tile(weather_annual, "sum_cold_noncrit", "PuBu", -1,
  "Accumulated non-critical cold  (vernalisation budget)",
  "Sum of t_dev for days where −15 ≤ t_dev < 0 °C — chilling requirement")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Temperature stress metrics by region and year",
    subtitle = paste0("Derived from daily Tmax / Tmin. Base temperature = ",
                      BASE_TEMP, " °C. Regions ordered cool → warm (W/N to E/S)."),
    caption  = "Source: ukraine_daily_weather.csv",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.subtitle   = element_text(size = 9, colour = "grey45"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
p_tmax <- make_tile(weather_annual, "mean_tmax", "RdYlBu", -1,
  "Mean daily Tmax  (°C)",
  "Annual average of daily maximum temperature")

p_tmin <- make_tile(weather_annual, "mean_tmin", "RdYlBu", -1,
  "Mean daily Tmin  (°C)",
  "Annual average of daily minimum temperature")

p_tmean <- make_tile(weather_annual, "mean_temp", "RdYlBu", -1,
  "Mean daily temperature  (°C)",
  "Annual average of (Tmax + Tmin) / 2")

(p_tmax | p_tmin | p_tmean) +
  plot_annotation(
    title   = "Raw temperature levels by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
make_tile(weather_annual, "total_precip", "BrBG", 1,
  "Total annual precipitation  (mm)",
  "Sum of daily precipitation. Green = wetter, brown = drier.") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
p_et0_mean <- make_tile(weather_annual, "mean_et0", "YlOrRd", 1,
  "Mean daily ET₀  (mm/day)",
  "Annual average reference evapotranspiration")

p_et0_total <- make_tile(weather_annual, "total_et0", "YlOrRd", 1,
  "Total annual ET₀  (mm/year)",
  "Sum of daily reference evapotranspiration")

(p_et0_mean | p_et0_total) +
  plot_annotation(
    title   = "Reference evapotranspiration (ET₀) by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
make_tile(weather_annual, "mean_wind", "PuRd", 1,
  "Mean annual maximum wind gusts  (m/s)",
  "Annual average of daily maximum wind gust at 10 m height") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
weather_annual |>
  group_by(region) |>
  summarise(
    `Hot days`     = round(mean(days_hot_crit),    1),
    `Cold days`    = round(mean(days_cold_crit),   1),
    `Heat sum`     = round(mean(sum_heat_noncrit), 0),
    `Cold sum`     = round(mean(sum_cold_noncrit), 0),
    `Precip (mm)`  = round(mean(total_precip),     0),
    `ET₀ (mm/yr)`  = round(mean(total_et0),        0),
    `Wind (m/s)`   = round(mean(mean_wind),        1)
  ) |>
  arrange(desc(`Hot days`)) |>
  knitr::kable(caption = "Mean annual weather metrics by oblast (2013–2024, ordered by heat stress days)")

```

### 2.3 Soil Data (`soil_data.csv`)

Static soil properties by oblast:

| Variable         | Description                        | Unit     |
|------------------|------------------------------------|----------|
| `oblast`         | Oblast name                        | text     |
| `pH`             | Soil acidity                       | unitless |
| `organic_matter` | Organic matter content             | \%       |
| `total_nitrogen` | Total nitrogen content             | \%       |
| `cec`            | Cation Exchange Capacity           | meq/100g |
| `clay`           | Clay content (\<0.002 mm)          | \%       |
| `sand`           | Sand content (0.05–2.0 mm)         | \%       |
| `silt`           | Silt content (0.002–0.05 mm)       | \%       |
| `ec`             | Electrical conductivity (salinity) | dS/m     |
| `esp`            | Exchangeable Sodium Percentage     | \%       |
| `c_n_ratio`      | Carbon to Nitrogen ratio           | unitless |

**Notes:** Optimal pH for winter wheat is 6.0–7.5. High EC (\>2 dS/m) indicates salinity stress. High ESP (\>15%) causes soil structure degradation. Low C:N ratio (\<15) favors rapid nitrogen release.

```{r}
# ── 13. Soil properties ──────────────────────────────────────
vars_soil <- c("pH", "organic_matter", "total_nitrogen",
               "cec", "clay", "sand", "silt")

p_soil <- soil |>
  pivot_longer(cols = all_of(vars_soil),
               names_to  = "property",
               values_to = "value") |>
  mutate(property = factor(property, levels = vars_soil)) |>
  ggplot(aes(x = value,
             y = reorder(oblast, value),
             fill = property)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  facet_wrap(~property, scales = "free_x", nrow = 2) +
  scale_fill_manual(
    values = c("#2c7bb6", "#d7191c", "#1a9641",
               "#fdae61", "#abd9e9", "#756bb1", "#a6bddb")
  ) +
  labs(
    title   = "Soil properties by oblast",
    caption = "Source: soil_data.csv"
  ) +
  theme_small_multiple(base_size = 8) +
  theme(
    axis.text.y    = element_text(size = 6),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3)
  )

# ── 14. Print all plots ──────────────────────────────────────

print(p_soil)
```

------------------------------------------------------------------------

# Methods.

Explain accurately what methods you are using to solve the tasks of interest.

To solve these tasks, we apply the following: 1. **Pooled OLS & Polynomial Regression:** To capture the parabolic effects of temperature and radiation. 2. **Segmented (Piecewise) Regression:** Specifically for our "Extra Days" hypothesis to model heat stress above $30^{\circ}C$. 3. **Fixed Effects (FE) Model:** To control for time-invariant regional factors like soil quality across Ukraine. 4. **Moderation Analysis:** Using interaction terms ($Yield \times Factor$) to see how natural factors alter the sensitivity of price to yield]. 5. **Correlation Matrix:** To test the $H_1$ of Hypothesis 4 and decide whether to omit cloudiness in favor of rain/shine to maintain model parsimony.

------------------------------------------------------------------------

# Results.

Describe the results you already have obtained at this stage.

At this stage (April 2, 2026), we have: - Completed the collection of meteorological data via Open-Meteo API and price data from Minfin. - Constructed a unified panel dataset aligned by weekly averages. - Initiated the classification of "Hot" and "Cold" days to test our Extreme Day hypothesis.

------------------------------------------------------------------------

# Conclusions.

What conclusions have you already drawn?

Early findings indicate that weather impacts are indeed non-linear, supporting our use of quadratic forms in the model. We have also concluded that high-frequency daily data is necessary to accurately count "extreme" events before aggregating them into the final econometric model.
