---
title: "Agrobusiness"
author: "Kate, Anastasiia, Max"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 7, dpi = 150)
```

------------------------------------------------------------------------

# Hypotheses and objectives description.

Our project researches the "story" of wheat market volatility in Ukraine. We are solving the problem of agricultural risk management by identifying which meteorological and human-led factors (like fertilizers) actually drive yields and prices. We aim to provide agribusinesses with a clear framework for data collection: what range of data is mandatory and what is redundant.

### Research Hypotheses

**1. Traditional Significance and Input Factors:**

$H_0$: Meteorological factors (radiation, temperature, humidity), fertilizers, and chemicals have no statistically significant effect on wheat yield or prices.

$H_1$: These factors have a significant impact, specifically following a non-linear "U-shaped" or parabolic curve where both extremes are detrimental.

**2. Seasonal Relevancy (The "Winter" Question):**

$H_0$: Weather data from outside the active spring/summer growing season (e.g., winter temperatures) is not significant for predicting harvest outcomes.

$H_1$: Winter weather is a significant predictor of soil moisture reserves and winter-kill rates, which directly impact the final yield and market price dynamics.

**3. Data Granularity and Extreme Day Classification:** - **Goal:** We test a custom classification where average daily temperature $(T_{max} - T_{min}) / 2$ is used to categorize days as "Extremely Cold" ($< -10^{\circ}C$), "Cold" ($-10^{\circ}C$ to $5^{\circ}C$), "Warm" ($5^{\circ}C$ to $30^{\circ}C$), or "Hot" ($> 30^{\circ}C$).

$H_0$: Summing "extra days" (extreme heat/cold) does not provide more predictive power than simple weekly averages.

$H_1$: The frequency of extreme temperature events is a more accurate predictor of yield shocks than mean temperature alone, requiring high-frequency (daily) data collection for valid modeling.

**4. Cloudiness and Correlation Logic (Variable Selection):**

$H_0$: Cloudiness, rain, and sunshine duration are independent predictors with no significant correlation.

$H_1$: Cloudiness is strongly correlated with precipitation and solar radiation. We hypothesize that using all three creates multicollinearity; therefore, we test which single variable (e.g., "shine") serves as the most effective proxy for the model.

------------------------------------------------------------------------

# Data analysis.

Provide your data analysis. What conclusions can be drawn from this analysis? Does this analysis lead to any first conclusions about the research question or suggest any hypotheses worth testing?

**Datasets:**

-   sowing_area.csv

-   harvested_area.csv

-   crop_capacity.csv

-   daily_weather.csv

-   soil_data.csv

------------------------------------------------------------------------

# Data Description and Exploratory Analysis

## 1. Introduction

This study investigates the impact of weather factors on winter wheat yield in Ukraine using panel data for 12 years (2013–2024) at the oblast level. The dependent variable is yield (centners per hectare). Key explanatory variables are derived from daily temperature data following a theoretically motivated aggregation scheme. Soil properties and area variables serve as controls.

------------------------------------------------------------------------

## 2. Data Description

This section describes all data sources used in the study: crop statistics (annual, oblast level), daily weather records, static soil properties, and derived weather metrics constructed following the research hypothesis.

### 2.1 Crop Data (annual, by oblast)

| File | Variable | Units | Explanation |
|------------------|------------------|------------------|------------------|
| `crop_capacity.csv` | Yield | centners/ha | Dependent variable. Already calculated by statistical authorities. |
| `harvested_area.csv` | Harvested area | hectares | Area from which crop was actually harvested. |
| `sowing_area.csv` | Sown area | hectares | Area planted (may differ from harvested due to losses). |

**Notes:** One centner = 100 kg. Typical winter wheat yield in Ukraine ranges from 30–80 ц/га. The difference between sown and harvested area indicates crop losses during vegetation.

```{r}
# ============================================================
# Interim Report – Part 1  (updated)
# Changes vs previous version:
#   1. Endpoint dots removed from all small-multiple plots
#   2. New section: harvested area as % of total oblast area
# ============================================================

library(tidyverse)
library(lubridate)
library(patchwork)

data_path <- "."
years     <- as.character(2015:2025)

META_COLS <- c("indicator", "region", "unit", "scale",
               "decimals", "special", "note_ts", "note_obs")

# ── Load helper ───────────────────────────────────────────────
read_panel <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names      = c(META_COLS, years),
    skip           = 1,
    show_col_types = FALSE
  ) |>
    select(region, all_of(years)) |>
    pivot_longer(cols = all_of(years), names_to = "year",
                 values_to = value_name) |>
    mutate(
      year         = as.integer(year),
      !!value_name := suppressWarnings(as.numeric(.data[[value_name]]))
    ) |>
    filter(region != "Україна")
}

read_ukraine <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names      = c(META_COLS, years),
    skip           = 1,
    show_col_types = FALSE
  ) |>
    filter(region == "Україна") |>
    select(region, all_of(years)) |>
    pivot_longer(cols = all_of(years), names_to = "year",
                 values_to = value_name) |>
    mutate(year = as.integer(year),
           !!value_name := suppressWarnings(as.numeric(.data[[value_name]])))
}

yield     <- read_panel("harvest_capacity.csv",  "yield_cpha")
harv      <- read_panel("harvested_area.csv",    "harv_kha") |>
               mutate(harv_ha = harv_kha * 1000) |> select(-harv_kha)
sown      <- read_panel("sowing_area.csv",       "sown_kha") |>
               mutate(sown_ha = sown_kha * 1000) |> select(-sown_kha)
ukr_yield <- read_ukraine("harvest_capacity.csv","yield_cpha")

# ── Region groups ─────────────────────────────────────────────
region_groups <- tribble(
  ~region,              ~group,
  "Волинська",          "West",
  "Закарпатська",       "West",
  "Івано-Франківська",  "West",
  "Львівська",          "West",
  "Рівненська",         "West",
  "Тернопільська",      "West",
  "Хмельницька",        "West",
  "Чернівецька",        "West",
  "Вінницька",          "Centre",
  "Дніпропетровська",   "Centre",
  "Житомирська",        "Centre",
  "Київська",           "Centre",
  "Кіровоградська",     "Centre",
  "Полтавська",         "Centre",
  "Сумська",            "Centre",
  "Харківська",         "Centre",
  "Черкаська",          "Centre",
  "Чернігівська",       "Centre",
  "Донецька",           "East",
  "Запорізька",         "East",
  "Луганська",          "East",
  "Миколаївська",       "East",
  "Одеська",            "East",
  "Херсонська",         "East"
)

GROUP_COLORS <- c(West = "#2c7bb6", Centre = "#d7191c", East = "#1a9641")

# ── Shared theme ──────────────────────────────────────────────
theme_small_multiple <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      panel.border      = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.line.x       = element_line(colour = "grey70", linewidth = 0.4),
      axis.ticks.x      = element_line(colour = "grey70", linewidth = 0.3),
      axis.ticks.y      = element_blank(),
      axis.text.x       = element_text(size = 7, colour = "grey40",
                                       margin = margin(t = 2)),
      axis.text.y       = element_text(size = 7, colour = "grey40"),
      axis.title        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8,
                                       colour = "grey20", margin = margin(b = 3)),
      strip.background  = element_blank(),
      strip.placement   = "outside",
      legend.position   = "none",
      plot.title        = element_text(face = "bold", size = 11,
                                       colour = "grey10", margin = margin(b = 4)),
      plot.subtitle     = element_text(size = 8.5, colour = "grey40",
                                       margin = margin(b = 8)),
      plot.caption      = element_text(size = 7, colour = "grey55",
                                       hjust = 0, margin = margin(t = 6)),
      plot.margin       = margin(10, 12, 8, 10)
    )
}

# ── Small-multiple helper  (NO endpoint dot) ──────────────────
plot_small_multiple <- function(df, group_map, value_col,
                                group    = NULL,
                                title    = "",
                                subtitle = "",
                                caption  = "") {

  df <- df |>
    left_join(group_map, by = "region") |>
    rename(value = !!sym(value_col))

  focal_regions <- if (!is.null(group)) {
    group_map |> filter(group == !!group) |> pull(region)
  } else {
    unique(df$region)
  }

  focal_color <- if (!is.null(group)) GROUP_COLORS[[group]] else "grey30"

  bg_data <- tibble(facet = focal_regions) |>
    cross_join(
      df |> filter(!is.na(value)) |> select(region, year, value)
    )

  focal_data <- df |>
    filter(region %in% focal_regions, !is.na(value)) |>
    mutate(facet = region)

  ggplot() +
    # grey background lines — all other regions
    geom_line(
      data    = bg_data |> filter(region != facet),
      mapping = aes(x = year, y = value, group = region),
      colour  = "grey85", linewidth = 0.35, alpha = 0.9
    ) +
    # coloured focal line — NO endpoint dot
    geom_line(
      data    = focal_data,
      mapping = aes(x = year, y = value, group = facet),
      colour  = focal_color, linewidth = 1.1, lineend = "round"
    ) +
    facet_wrap(~facet, scales = "free_y") +
    scale_x_continuous(
      breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
      labels = c("'15", "'17", "'19", "'21", "'23", "'25")
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.10))) +
    labs(title = title, subtitle = subtitle, caption = caption) +
    theme_small_multiple()
}

# ── Yield plots ───────────────────────────────────────────────
p_yield_west <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "West",
  title    = "Winter wheat yield — Western oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_yield_centre <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "Centre",
  title    = "Winter wheat yield — Central oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_yield_east <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "East",
  title    = "Winter wheat yield — Eastern oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── Harvested area plots ──────────────────────────────────────
p_harv_west <- plot_small_multiple(
  harv, region_groups, "harv_ha", "West",
  title    = "Harvested area — Western oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_harv_centre <- plot_small_multiple(
  harv, region_groups, "harv_ha", "Centre",
  title    = "Harvested area — Central oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_harv_east <- plot_small_multiple(
  harv, region_groups, "harv_ha", "East",
  title    = "Harvested area — Eastern oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── Sown area plots ───────────────────────────────────────────
p_sown_west <- plot_small_multiple(
  sown, region_groups, "sown_ha", "West",
  title    = "Sown area — Western oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_sown_centre <- plot_small_multiple(
  sown, region_groups, "sown_ha", "Centre",
  title    = "Sown area — Central oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)
p_sown_east <- plot_small_multiple(
  sown, region_groups, "sown_ha", "East",
  title    = "Sown area — Eastern oblasts",
  subtitle = "Hectares, 2015–2025.",
  caption  = "Source: State Statistics Service of Ukraine"
)

# ── Representative oblasts ────────────────────────────────────
rep_oblasts <- yield |>
  left_join(region_groups, by = "region") |>
  filter(!is.na(group), !is.na(yield_cpha)) |>
  group_by(group, region) |>
  summarise(mean_y = mean(yield_cpha, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_y, n = 1, by = group)

rep_data <- yield |>
  inner_join(rep_oblasts |> select(group, region), by = "region") |>
  filter(!is.na(yield_cpha))

p_representative <- ggplot() +
  geom_line(
    data    = ukr_yield |> filter(!is.na(yield_cpha)),
    mapping = aes(x = year, y = yield_cpha),
    colour  = "grey65", linewidth = 0.7, linetype = "dashed"
  ) +
  geom_line(
    data    = rep_data,
    mapping = aes(x = year, y = yield_cpha, colour = group),
    linewidth = 1.2, lineend = "round"
  ) +
  facet_wrap(~group) +
  scale_colour_manual(values = GROUP_COLORS) +
  scale_x_continuous(
    breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
    labels = c("'15", "'17", "'19", "'21", "'23", "'25")
  ) +
  labs(
    title    = "Most productive representative oblast per geographical group",
    subtitle = "Dashed grey = Ukraine national average. Coloured = representative oblast (highest mean yield).",
    caption  = "Source: State Statistics Service of Ukraine"
  ) +
  theme_small_multiple(base_size = 10) +
  theme(strip.text = element_text(face = "bold", size = 10, colour = "grey15"))

# ════════════════════════════════════════════════════════════
# HARVESTED AREA AS % OF TOTAL OBLAST AREA
# ════════════════════════════════════════════════════════════
#
# Total oblast areas (km²) from official Ukrainian sources.
# Converted to hectares (1 km² = 100 ha).
# These are fixed geographic areas — they do not change year to year.

oblast_area_km2 <- tribble(
  ~region,              ~total_km2,
  "Вінницька",          26513,
  "Волинська",          20144,
  "Дніпропетровська",   31914,
  "Донецька",           26517,
  "Житомирська",        29832,
  "Закарпатська",       12777,
  "Запорізька",         27180,
  "Івано-Франківська",  13900,
  "Київська",           28131,
  "Кіровоградська",     24588,
  "Луганська",          26683,
  "Львівська",          21833,
  "Миколаївська",       24598,
  "Одеська",            33310,
  "Полтавська",         28748,
  "Рівненська",         20047,
  "Сумська",            23832,
  "Тернопільська",      13823,
  "Харківська",         31415,
  "Херсонська",         28461,
  "Хмельницька",        20645,
  "Черкаська",          20900,
  "Чернівецька",         8097,
  "Чернігівська",       31865
) |>
  mutate(total_ha = total_km2 * 100)

# Join and compute percentage
harv_pct <- harv |>
  left_join(oblast_area_km2 |> select(region, total_ha), by = "region") |>
  mutate(harv_pct = harv_ha / total_ha * 100) |>
  filter(!is.na(harv_pct))

# Quick sanity check — print min/max values
harv_pct |>
  group_by(region) |>
  summarise(mean_pct = round(mean(harv_pct, na.rm = TRUE), 1),
            min_pct  = round(min(harv_pct,  na.rm = TRUE), 1),
            max_pct  = round(max(harv_pct,  na.rm = TRUE), 1)) |>
  arrange(desc(mean_pct)) |>
  print(n = 30)

# ── % plots (same small-multiple helper, new variable) ────────
p_pct_west <- plot_small_multiple(
  harv_pct, region_groups, "harv_pct", "West",
  title    = "Harvested wheat area as % of total oblast area — West",
  subtitle = "% of total oblast land area under winter wheat harvest, 2015–2025.",
  caption  = "Oblast areas: official Ukrainian administrative data. Harvested area: SSU."
)
p_pct_centre <- plot_small_multiple(
  harv_pct, region_groups, "harv_pct", "Centre",
  title    = "Harvested wheat area as % of total oblast area — Centre",
  subtitle = "% of total oblast land area under winter wheat harvest, 2015–2025.",
  caption  = "Oblast areas: official Ukrainian administrative data. Harvested area: SSU."
)
p_pct_east <- plot_small_multiple(
  harv_pct, region_groups, "harv_pct", "East",
  title    = "Harvested wheat area as % of total oblast area — East",
  subtitle = "% of total oblast land area under winter wheat harvest, 2015–2025.",
  caption  = "Oblast areas: official Ukrainian administrative data. Harvested area: SSU."
)

# ── Single cross-group summary: % intensity bar chart ─────────
# Average % across all years per oblast — useful for one-shot comparison
harv_pct_avg <- harv_pct |>
  left_join(region_groups, by = "region") |>
  group_by(region, group) |>
  summarise(mean_pct = mean(harv_pct, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_pct) |>
  mutate(region = factor(region, levels = region))

p_pct_bar <- ggplot(harv_pct_avg,
                    aes(x = mean_pct, y = region, fill = group)) +
  geom_col(width = 0.72, alpha = 0.88) +
  geom_text(aes(label = paste0(round(mean_pct, 1), "%")),
            hjust = -0.1, size = 2.6, colour = "grey30") +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::label_number(suffix = "%")
  ) +
  labs(
    title    = "Average harvested wheat area as % of total oblast area (2015–2025)",
    subtitle = "Oblasts ordered by intensity. Colour = geographical group.",
    caption  = "Oblast areas: official Ukrainian administrative data. Harvested area: SSU.",
    fill     = "Group"
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    axis.text.y        = element_text(size = 8),
    legend.position    = "right",
    legend.text        = element_text(size = 8),
    legend.title       = element_text(size = 8, face = "bold")
  )

# ── Print all ─────────────────────────────────────────────────
print(p_yield_west)
print(p_yield_centre)
print(p_yield_east)
print(p_harv_west)
print(p_harv_centre)
print(p_harv_east)
print(p_sown_west)
print(p_sown_centre)
print(p_sown_east)
print(p_representative)
print(p_pct_west)
print(p_pct_centre)
print(p_pct_east)
print(p_pct_bar)

# ── Save (uncomment as needed) ────────────────────────────────
# ggsave("yield_west.png",      p_yield_west,    width=12, height=8, dpi=300, bg="white")
# ggsave("yield_centre.png",    p_yield_centre,  width=12, height=8, dpi=300, bg="white")
# ggsave("yield_east.png",      p_yield_east,    width=12, height=8, dpi=300, bg="white")
# ggsave("harv_pct_west.png",   p_pct_west,      width=12, height=8, dpi=300, bg="white")
# ggsave("harv_pct_centre.png", p_pct_centre,    width=12, height=8, dpi=300, bg="white")
# ggsave("harv_pct_east.png",   p_pct_east,      width=12, height=8, dpi=300, bg="white")
# ggsave("harv_pct_bar.png",    p_pct_bar,       width=10, height=8, dpi=300, bg="white")
# ggsave("representative.png",  p_representative,width=12, height=5, dpi=300, bg="white")

```

### 2.2 Daily Weather Data (`daily_weather.csv`)

| Variable | Description | Unit |
|------------------------|------------------------|------------------------|
| `region` | Oblast name | text |
| `date` | Calendar date | YYYY-MM-DD |
| `temperature_2m_max` | Daily maximum temperature | °C |
| `temperature_2m_min` | Daily minimum temperature | °C |
| `temperature_2m_mean` | Daily mean temperature | °C |
| `wind_speed_10m_max` | Daily maximum wind speed at 10m height | m/s |
| `wind_gusts_10m_max` | Daily maximum wind gusts at 10m height | m/s |
| `et0_fao_evapotranspiration` | Reference evapotranspiration (FAO Penman-Monteith) | mm/day |
| `dewpoint_2m_mean` | Daily mean dewpoint temperature | °C |
| `precipitation_sum` | Daily total precipitation | mm |
| `shortwave_radiation_sum` | Daily incoming shortwave solar radiation | MJ/m² |

**Notes:** ET0 represents atmospheric evaporative demand — higher values indicate faster water loss from crops. Dewpoint is a direct measure of air moisture (higher values promote fungal diseases). Data cover 2013–2024 for all 24 oblasts.

#### Constructed Weather Metrics (following the research hypothesis)

For each region and year, daily temperature is transformed as:

$$ t_{dev} = \frac{T_{max} + T_{min}}{2} - 5 $$

where 5°C is the assumed base temperature for winter wheat development.

From this deviation, four metrics are calculated for the **full year** (later the vegetation period will be tested):

| Metric | Definition |
|------------------------------------|------------------------------------|
| `days_hot_critical` | Number of days with $t_{dev} > 20$ (extreme heat stress) |
| `days_cold_critical` | Number of days with $t_{dev} < -15$ (extreme cold stress) |
| `sum_heat_noncrit` | Sum of $t_{dev}$ for days with $0 < t_{dev} \leq 20$ |
| `sum_cold_noncrit` | Sum of $t_{dev}$ for days with $-15 \leq t_{dev} < 0$ |

**Notes:** These variables capture different aspects of temperature stress: critical extremes (counts) and accumulated thermal resources (sums). Thresholds (20°C, -15°C) are based on agronomic literature for winter wheat.

```{r fig.width=14, fig.height=9}
# ── Load and aggregate ─────────────────────────────────────
weather <- read_csv("daily_weather.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date), year = year(date)) |>
  filter(year >= 2013, year <= 2024)

BASE_TEMP <- 5

weather_annual <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - BASE_TEMP,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20,  t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0,  t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    days_hot_crit    = sum(days_hot_crit,  na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit, na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,   na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,   na.rm = TRUE),
    mean_tmax        = mean(temperature_2m_max,          na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min,          na.rm = TRUE),
    mean_temp        = mean(t_mean,                      na.rm = TRUE),
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    mean_et0         = mean(et0_fao_evapotranspiration,  na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    total_radiation = sum(shortwave_radiation_sum, na.rm = TRUE),
    .groups = "drop"
  )

# Order regions W→E by mean temperature (warmer = further east/south)
region_order <- weather_annual |>
  group_by(region) |>
  summarise(mt = mean(mean_temp)) |>
  arrange(mt) |>
  pull(region)

weather_annual <- weather_annual |>
  mutate(region = factor(region, levels = region_order),
         year   = factor(year))
# ── Shared theme ────────────────────────────────────────────
theme_heatmap <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid        = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.text.x       = element_text(size = 7.5, angle = 45, hjust = 1,
                                       colour = "grey35"),
      axis.text.y       = element_text(size = 7,   colour = "grey30"),
      axis.title        = element_blank(),
      axis.ticks        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8.5,
                                       colour = "grey15"),
      strip.background  = element_blank(),
      plot.title        = element_text(face = "bold", size = 10,
                                       colour = "grey10",
                                       margin = margin(b = 2)),
      plot.subtitle     = element_text(size = 7.5, colour = "grey45",
                                       margin = margin(b = 5)),
      legend.text       = element_text(size = 7,   colour = "grey35"),
      legend.title      = element_blank(),
      legend.key.width  = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin       = margin(6, 8, 6, 6)
    )
}

# ── Helper: one heatmap tile ────────────────────────────────
make_tile <- function(df, metric, palette, direction = 1,
                      title, subtitle) {
  ggplot(df, aes(x = year, y = region, fill = .data[[metric]])) +
    geom_tile(colour = "white", linewidth = 0.35) +
    scale_fill_distiller(palette   = palette,
                         direction = direction,
                         guide     = guide_colorbar(
                           barwidth  = 0.45,
                           barheight = 7,
                           ticks     = FALSE)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(title = title, subtitle = subtitle) +
    theme_heatmap()
}
p1 <- make_tile(weather_annual, "days_hot_crit", "YlOrRd", 1,
  "Days with extreme heat stress",
  "t_dev > 20 °C  (mean daily temp > 25 °C) — count of days per year")

p2 <- make_tile(weather_annual, "days_cold_crit", "Blues", -1,
  "Days with extreme cold stress",
  "t_dev < −15 °C  (mean daily temp < −10 °C) — count of days per year")

p3 <- make_tile(weather_annual, "sum_heat_noncrit", "YlOrBr", 1,
  "Accumulated non-critical heat  (growing degree-days)",
  "Sum of t_dev for days where 0 < t_dev ≤ 20 °C — thermal budget for growth")

p4 <- make_tile(weather_annual, "sum_cold_noncrit", "PuBu", -1,
  "Accumulated non-critical cold  (vernalisation budget)",
  "Sum of t_dev for days where −15 ≤ t_dev < 0 °C — chilling requirement")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Temperature stress metrics by region and year",
    subtitle = paste0("Derived from daily Tmax / Tmin. Base temperature = ",
                      BASE_TEMP, " °C. Regions ordered cool → warm (W/N to E/S)."),
    caption  = "Source: ukraine_daily_weather.csv",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.subtitle   = element_text(size = 9, colour = "grey45"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
p_tmax <- make_tile(weather_annual, "mean_tmax", "RdYlBu", -1,
  "Mean daily Tmax  (°C)",
  "Annual average of daily maximum temperature")

p_tmin <- make_tile(weather_annual, "mean_tmin", "RdYlBu", -1,
  "Mean daily Tmin  (°C)",
  "Annual average of daily minimum temperature")

p_tmean <- make_tile(weather_annual, "mean_temp", "RdYlBu", -1,
  "Mean daily temperature  (°C)",
  "Annual average of (Tmax + Tmin) / 2")

(p_tmax | p_tmin | p_tmean) +
  plot_annotation(
    title   = "Raw temperature levels by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
make_tile(weather_annual, "total_precip", "BrBG", 1,
  "Total annual precipitation  (mm)",
  "Sum of daily precipitation. Green = wetter, brown = drier.") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
make_tile(weather_annual, "total_et0", "YlOrRd", 1,
  "Total annual ET₀  (mm/year)",
  "Sum of daily reference evapotranspiration") +
  plot_annotation(
    title   = "Reference evapotranspiration (ET₀) by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
make_tile(weather_annual, "mean_wind", "PuRd", 1,
  "Mean annual maximum wind gusts  (m/s)",
  "Annual average of daily maximum wind gust at 10 m height") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
make_tile(weather_annual, "total_radiation", "YlOrRd", 1,
  "Total annual solar radiation  (MJ/m²)",
  "Sum of daily shortwave radiation") +
  plot_annotation(
    title   = "Solar radiation by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
weather_annual |>
  group_by(region) |>
  summarise(
    `Hot days`     = round(mean(days_hot_crit),    1),
    `Cold days`    = round(mean(days_cold_crit),   1),
    `Heat sum`     = round(mean(sum_heat_noncrit), 0),
    `Cold sum`     = round(mean(sum_cold_noncrit), 0),
    `Precip (mm)`  = round(mean(total_precip),     0),
    `ET₀ (mm/yr)`  = round(mean(total_et0),        0),
    `Wind (m/s)`   = round(mean(mean_wind),        1)
  ) |>
  arrange(desc(`Hot days`)) |>
  knitr::kable(caption = "Mean annual weather metrics by oblast (2013–2024, ordered by heat stress days)")

```

### 2.3 Soil Data (`soil_data.csv`)

Static soil properties by oblast:

| Variable         | Description                        | Unit     |
|------------------|------------------------------------|----------|
| `oblast`         | Oblast name                        | text     |
| `pH`             | Soil acidity                       | unitless |
| `organic_matter` | Organic matter content             | \%       |
| `total_nitrogen` | Total nitrogen content             | \%       |
| `cec`            | Cation Exchange Capacity           | meq/100g |
| `clay`           | Clay content (\<0.002 mm)          | \%       |
| `sand`           | Sand content (0.05–2.0 mm)         | \%       |
| `silt`           | Silt content (0.002–0.05 mm)       | \%       |
| `ec`             | Electrical conductivity (salinity) | dS/m     |
| `esp`            | Exchangeable Sodium Percentage     | \%       |
| `c_n_ratio`      | Carbon to Nitrogen ratio           | unitless |

**Notes:** Optimal pH for winter wheat is 6.0–7.5. High EC (\>2 dS/m) indicates salinity stress. High ESP (\>15%) causes soil structure degradation. Low C:N ratio (\<15) favors rapid nitrogen release.

```{r}
# ── 13. Soil properties ──────────────────────────────────────
vars_soil <- c("pH", "organic_matter", "total_nitrogen",
               "cec", "clay", "sand", "silt")

p_soil <- soil |>
  pivot_longer(cols = all_of(vars_soil),
               names_to  = "property",
               values_to = "value") |>
  mutate(property = factor(property, levels = vars_soil)) |>
  ggplot(aes(x = value,
             y = reorder(oblast, value),
             fill = property)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  facet_wrap(~property, scales = "free_x", nrow = 2) +
  scale_fill_manual(
    values = c("#2c7bb6", "#d7191c", "#1a9641",
               "#fdae61", "#abd9e9", "#756bb1", "#a6bddb")
  ) +
  labs(
    title   = "Soil properties by oblast",
    caption = "Source: soil_data.csv"
  ) +
  theme_small_multiple(base_size = 8) +
  theme(
    axis.text.y    = element_text(size = 6),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3)
  )

# ── 14. Print all plots ──────────────────────────────────────

print(p_soil)
```

### 2.4 Correlation structure of all variables

In exploratory research it is important to check correlations not only among variables we expect to be related, but across all pairs — including crop variables, area variables, soil properties, and weather metrics together. Unexpected correlations can reveal data artefacts, suggest omitted mechanisms, or warn about multicollinearity before model estimation.

------------------------------------------------------------------------

### Setup: build the full panel

```{r build-full-panel}
library(tidyverse)
library(corrplot)
library(GGally)
library(car)

# ── Region name lookup (Latin weather ↔ Cyrillic crop data) ──
# Adjust spellings to match exactly what appears in your files
region_lookup <- tribble(
  ~region_lat,        ~region_cyr,
  "Vinnytsia",        "Вінницька",
  "Volyn",            "Волинська",
  "Dnipropetrovsk",   "Дніпропетровська",
  "Donetsk",          "Донецька",
  "Zhytomyr",         "Житомирська",
  "Zakarpattia",      "Закарпатська",
  "Zaporizhzhia",     "Запорізька",
  "Ivano-Frankivsk",  "Івано-Франківська",
  "Kyiv",             "Київська",
  "Kirovohrad",       "Кіровоградська",
  "Luhansk",          "Луганська",
  "Lviv",             "Львівська",
  "Mykolaiv",         "Миколаївська",
  "Odesa",            "Одеська",
  "Poltava",          "Полтавська",
  "Rivne",            "Рівненська",
  "Sumy",             "Сумська",
  "Ternopil",         "Тернопільська",
  "Kharkiv",          "Харківська",
  "Kherson",          "Херсонська",
  "Khmelnytskyi",     "Хмельницька",
  "Cherkasy",         "Черкаська",
  "Chernivtsi",       "Чернівецька",
  "Chernihiv",        "Чернігівська"
)

# ── Weather annual aggregation (all variables) ────────────────
weather_annual_full <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - 5,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20, t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0, t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    # Thermal stress metrics
    days_hot_crit    = sum(days_hot_crit,       na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit,      na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,        na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,        na.rm = TRUE),
    # Raw temperature levels
    mean_tmax        = mean(temperature_2m_max, na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min, na.rm = TRUE),
    mean_temp        = mean(t_mean,             na.rm = TRUE),
    # Other weather
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    total_radiation  = sum(shortwave_radiation_sum,      na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    mean_dewpoint    = mean(dewpoint_2m_mean,            na.rm = TRUE),
    # Derived
    water_deficit    = sum(precipitation_sum, na.rm = TRUE) -
                       sum(et0_fao_evapotranspiration, na.rm = TRUE),
    diurnal_range    = mean(temperature_2m_max - temperature_2m_min,
                            na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(year >= 2015)

# ── Harvest ratio ─────────────────────────────────────────────
harv_ratio_df <- harv |>
  left_join(sown, by = c("region", "year")) |>
  mutate(harv_ratio = pmin(harv_ha / sown_ha, 1.05)) |>
  select(region, year, harv_ha, sown_ha, harv_ratio)

# ── Join everything on Cyrillic region names ──────────────────
panel_full <- yield |>
  left_join(harv_ratio_df,  by = c("region", "year")) |>
  left_join(region_lookup,  by = c("region" = "region_cyr")) |>
  left_join(weather_annual_full,
            by = c("region_lat" = "region", "year")) |>
  left_join(soil |> rename(region_cyr = oblast) |>
              left_join(region_lookup, by = c("region_cyr" = "region_cyr")),
            by = c("region" = "region_cyr")) |>
  filter(!is.na(yield_cpha))

cat("Panel rows:", nrow(panel_full), "\n")
cat("Regions matched:", n_distinct(panel_full$region[!is.na(panel_full$mean_temp)]), "\n")
```

```{r collinearity, fig.width = 10, fig.height = 9}
# ── Regressor set: only variables entering the model ──────────
# Four temperature metrics + raw weather + harvest ratio
# Soil variables are time-invariant and will be absorbed by oblast
# fixed effects in the panel model, so they are excluded here.

regressors <- c(
  "days_hot_crit",
  "days_cold_crit",
  "sum_heat_noncrit",
  "sum_cold_noncrit",
  "total_precip",
  "total_et0",
  "total_radiation",
  "mean_wind",
  "mean_dewpoint",
  "harv_ratio"
)

# keep only columns that exist
regressors <- intersect(regressors, names(panel_full))

reg_df <- panel_full |>
  select(all_of(regressors)) |>
  drop_na()


# ── Display labels ────────────────────────────────────────────
reg_labels <- c(
  days_hot_crit    = "Hot critical days",
  days_cold_crit   = "Cold critical days",
  sum_heat_noncrit = "Accum. heat",
  sum_cold_noncrit = "Accum. cold",
  total_precip     = "Precipitation",
  total_et0        = "ET0",
  total_radiation  = "Solar radiation",
  mean_wind        = "Wind gusts",
  mean_dewpoint    = "Dewpoint",
  harv_ratio       = "Harvest ratio"
)

reg_labels <- reg_labels[regressors]

# ── 1. Correlation matrix among regressors ────────────────────
cor_reg <- cor(reg_df, use = "pairwise.complete.obs")
rownames(cor_reg) <- colnames(cor_reg) <- reg_labels

par(bg = "white")
corrplot(
  cor_reg,
  method      = "color",
  type        = "upper",
  order       = "hclust",
  tl.col      = "grey15",
  tl.srt      = 40,
  tl.cex      = 0.85,
  cl.cex      = 0.75,
  addCoef.col = "grey10",
  number.cex  = 0.72,
  col         = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(200),
  title       = "Correlation matrix among regressors",
  mar         = c(0, 0, 2, 0)
)
```

```{r vif, fig.width = 7, fig.height = 5}
# ── 2. VIF for the full regressor set ─────────────────────────
# OLS without fixed effects — fixed effects are not needed here
# because VIF measures regressor redundancy, not causal identification.
# Adding a constant outcome variable (yield) is enough.

vif_formula <- as.formula(
  paste("yield_cpha ~", paste(regressors, collapse = " + "))
)

vif_data <- panel_full |>
  select(yield_cpha, all_of(regressors)) |>
  drop_na()

vif_model  <- lm(vif_formula, data = vif_data)
vif_values <- vif(vif_model)

vif_df <- tibble(
  variable = names(vif_values),
  VIF      = vif_values
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High — must address",
      VIF > 5  ~ "Moderate — monitor",
      TRUE     ~ "Acceptable"
    ),
    concern = factor(concern,
                     levels = c("Acceptable",
                                "Moderate — monitor",
                                "High — must address"))
  ) |>
  arrange(desc(VIF))

# table
knitr::kable(
  vif_df |> select(label, VIF, concern),
  digits  = 2,
  col.names = c("Variable", "VIF", "Assessment"),
  caption = "Variance Inflation Factors — full regressor set"
)
```

```{r vif-plot, fig.width = 8, fig.height = 5}
# VIF bar chart — easier to read than a table for the report
ggplot(vif_df,
       aes(x = VIF,
           y = fct_reorder(label, VIF),
           fill = concern)) +
  geom_col(width = 0.68, alpha = 0.88) +
  geom_text(aes(label = round(VIF, 1)),
            hjust = -0.15, size = 3, colour = "grey25") +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#d7191c", linewidth = 0.5) +
  annotate("text", x = 5.2,  y = 0.6, label = "VIF = 5",
           hjust = 0, size = 2.6, colour = "grey50") +
  annotate("text", x = 10.2, y = 0.6, label = "VIF = 10",
           hjust = 0, size = 2.6, colour = "#d7191c") +
  scale_fill_manual(values = c(
    "Acceptable"           = "#1a9641",
    "Moderate — monitor"   = "#fdae61",
    "High — must address"  = "#d7191c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title   = "Variance Inflation Factors — full regressor set",
    x       = "VIF", y = NULL, fill = NULL,
    caption = "Dashed grey line = VIF 5 threshold. Dashed red line = VIF 10 threshold."
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    legend.position    = "top"
  )
```

```{r vif-reduced, fig.width = 8, fig.height = 5}
# ── 3. VIF after dropping the most collinear variables ─────────
# If ET0 and solar radiation both produce high VIF, drop solar radiation
# (absorbed by ET0) and recheck. This iterative approach shows clearly
# which variable is causing the inflation.

regressors_reduced <- regressors[!regressors %in% c("total_radiation",
                                                      "mean_dewpoint")]

vif_formula2 <- as.formula(
  paste("yield_cpha ~", paste(regressors_reduced, collapse = " + "))
)

vif_model2  <- lm(vif_formula2,
                  data = panel_full |>
                    select(yield_cpha, all_of(regressors_reduced)) |>
                    drop_na())
vif_values2 <- vif(vif_model2)

vif_df2 <- tibble(
  variable = names(vif_values2),
  VIF      = vif_values2
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High — must address",
      VIF > 5  ~ "Moderate — monitor",
      TRUE     ~ "Acceptable"
    ),
    concern = factor(concern,
                     levels = c("Acceptable",
                                "Moderate — monitor",
                                "High — must address"))
  ) |>
  arrange(desc(VIF))

ggplot(vif_df2,
       aes(x = VIF,
           y = fct_reorder(label, VIF),
           fill = concern)) +
  geom_col(width = 0.68, alpha = 0.88) +
  geom_text(aes(label = round(VIF, 1)),
            hjust = -0.15, size = 3, colour = "grey25") +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#d7191c", linewidth = 0.5) +
  annotate("text", x = 5.2,  y = 0.6, label = "VIF = 5",
           hjust = 0, size = 2.6, colour = "grey50") +
  annotate("text", x = 10.2, y = 0.6, label = "VIF = 10",
           hjust = 0, size = 2.6, colour = "#d7191c") +
  scale_fill_manual(values = c(
    "Acceptable"           = "#1a9641",
    "Moderate — monitor"   = "#fdae61",
    "High — must address"  = "#d7191c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title   = "VIF after removing solar radiation and dewpoint",
    subtitle = "Shows whether removing the most collinear variables resolves inflation in the remaining set.",
    x       = "VIF", y = NULL, fill = NULL,
    caption = "Dashed grey line = VIF 5. Dashed red line = VIF 10."
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    legend.position    = "top"
  )
```

------------------------------------------------------------------------

# 3. Modelling

This section builds a sequence of nested two-way fixed-effects panel models that together address the four hypotheses stated in the introduction. All models use the **within estimator** (region FE + year FE) estimated with the `fixest` package, which handles unbalanced panels and clustered standard errors efficiently. Soil properties are time-invariant within regions and are therefore fully absorbed by region fixed effects — they do not enter the regression formulas explicitly. Model fit is compared using AIC, BIC, overall R², and within-R² (variation explained after removing region and year means).

------------------------------------------------------------------------

## 3.1 Name harmonisation and panel construction

The three data sources use three different region spellings. We extend the existing `region_lookup` table with a `region_soil` column that maps the soil-file names ("Vinnytska") to the same spine, so every join goes through a single lookup and no manual string-fixing is needed elsewhere.

```{r model-name-lookup}
# ── Complete region lookup table (one source of truth) ─────────
# This extends the previously defined region_lookup with soil-file spellings.
# If region_lookup doesn't exist yet (e.g., running chunks out of order),
# we create it from scratch.

if (!exists("region_lookup")) {
  # Create base lookup if not already defined
  region_lookup <- tribble(
    ~region_lat,        ~region_cyr,
    "Vinnytsia",        "Вінницька",
    "Volyn",            "Волинська",
    "Dnipropetrovsk",   "Дніпропетровська",
    "Donetsk",          "Донецька",
    "Zhytomyr",         "Житомирська",
    "Zakarpattia",      "Закарпатська",
    "Zaporizhzhia",     "Запорізька",
    "Ivano-Frankivsk",  "Івано-Франківська",
    "Kyiv",             "Київська",
    "Kirovohrad",       "Кіровоградська",
    "Luhansk",          "Луганська",
    "Lviv",             "Львівська",
    "Mykolaiv",         "Миколаївська",
    "Odesa",            "Одеська",
    "Poltava",          "Полтавська",
    "Rivne",            "Рівненська",
    "Sumy",             "Сумська",
    "Ternopil",         "Тернопільська",
    "Kharkiv",          "Харківська",
    "Kherson",          "Херсонська",
    "Khmelnytskyi",     "Хмельницька",
    "Cherkasy",         "Черкаська",
    "Chernivtsi",       "Чернівецька",
    "Chernihiv",        "Чернігівська"
  )
}

# Add soil-file spellings (Latin, -ska suffix)
region_lookup <- region_lookup |>
  mutate(
    region_soil = case_match(
      region_lat,
      "Vinnytsia"       ~ "Vinnytska",
      "Volyn"           ~ "Volynska",
      "Dnipropetrovsk"  ~ "Dnipropetrovska",
      "Donetsk"         ~ "Donetska",
      "Zhytomyr"        ~ "Zhytomyrska",
      "Zakarpattia"     ~ "Zakarpatska",
      "Zaporizhzhia"    ~ "Zaporizka",
      "Ivano-Frankivsk" ~ "Ivano-Frankivska",
      "Kyiv"            ~ "Kyivska",
      "Kirovohrad"      ~ "Kirovohradska",
      "Luhansk"         ~ "Luhanska",
      "Lviv"            ~ "Lvivska",
      "Mykolaiv"        ~ "Mykolaivska",
      "Odesa"           ~ "Odeska",
      "Poltava"         ~ "Poltavska",
      "Rivne"           ~ "Rivnenska",
      "Sumy"            ~ "Sumska",
      "Ternopil"        ~ "Ternopilska",
      "Kharkiv"         ~ "Kharkivska",
      "Kherson"         ~ "Khersonska",
      "Khmelnytskyi"    ~ "Khmelnytska",
      "Cherkasy"        ~ "Cherkaska",
      "Chernivtsi"      ~ "Chernivetska",
      "Chernihiv"       ~ "Chernihivska"
    )
  )

# Verify — every row should have all three name columns filled
stopifnot(all(!is.na(region_lookup$region_soil)))

# Optional: show what we have
cat("Region lookup complete:", nrow(region_lookup), "oblasts\n")
```

------------------------------------------------------------------------

## 3.2 Growing-season weather aggregates

Winter wheat in Ukraine follows a well-documented phenological calendar:

| Phase | Calendar window | Biological meaning |
|------------------|------------------------|------------------------------|
| Pre-sowing buffer | 10 Sep – 30 Sep | Soil moisture build-up before sowing |
| Sowing → dormancy (**early half**) | 1 Oct – 15 May | Vernalisation, winter dormancy, tillering, green-up |
| Anthesis → harvest (**late half**) | 16 May – 31 Jul | Heading, grain filling, maturity |
| Full growing season | 10 Sep – 31 Jul | All of the above combined |

The split at **15 May** is the biologically meaningful boundary between the cold-sensitive vegetative phase and the heat-sensitive reproductive phase. Heat stress during anthesis (Feekes 10.5) has a particularly strong negative effect on grain set, making late-half temperature metrics critical predictors (Hypothesis 3).

```{r season-aggregates}
library(lubridate)  # ensure day(), month(), year() are available

BASE_TEMP <- 5   # base temperature for winter wheat development

# ── Shared aggregation helper ─────────────────────────────────
# Applies t_dev logic on any pre-filtered slice of daily data.
# Returns annual aggregates by region for the specified time window.

agg_weather_slice <- function(df) {
  df |>
    mutate(
      t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
      t_dev          = t_mean - BASE_TEMP,
      days_hot_crit  = as.integer(t_dev > 20),
      days_cold_crit = as.integer(t_dev < -15),
      heat_noncrit   = if_else(t_dev > 0 & t_dev <= 20, t_dev, 0),
      cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0, t_dev, 0)
    ) |>
    group_by(region, year) |>
    summarise(
      days_hot_crit    = sum(days_hot_crit,               na.rm = TRUE),
      days_cold_crit   = sum(days_cold_crit,              na.rm = TRUE),
      sum_heat_noncrit = sum(heat_noncrit,                na.rm = TRUE),
      sum_cold_noncrit = sum(cold_noncrit,                na.rm = TRUE),
      total_precip     = sum(precipitation_sum,           na.rm = TRUE),
      total_et0        = sum(et0_fao_evapotranspiration,  na.rm = TRUE),
      mean_wind        = mean(wind_gusts_10m_max,         na.rm = TRUE),
      water_deficit    = sum(precipitation_sum,           na.rm = TRUE) - 
                         sum(et0_fao_evapotranspiration,  na.rm = TRUE),
      .groups = "drop"
    )
}

# ── Reassign days to harvest year ─────────────────────────────
# Growing season spans September (year Y) to July (year Y+1).
# Sep–Dec of year Y  → harvest year Y+1
# Jan–Jul of year Y+1 → harvest year Y+1

weather_tagged <- weather |>
  mutate(
    cal_month  = month(date),
    cal_day    = day(date),
    cal_year   = year(date),
    harvest_yr = if_else(cal_month >= 9, cal_year + 1, cal_year)
  ) |>
  select(-year)   # remove original calendar-year column

# ── Full growing season: 10 Sep – 31 Jul ──────────────────────
weather_gs <- weather_tagged |>
  filter(
    (cal_month == 9 & cal_day >= 10) |   # from 10 Sep
    (cal_month %in% c(10, 11, 12)) |     # Oct–Dec
    (cal_month %in% c(1, 2, 3, 4, 5, 6, 7))  # Jan–Jul (full months)
  ) |>
  rename(year = harvest_yr) |>
  agg_weather_slice() |>
  rename_with(~ paste0(.x, "_gs"), -c(region, year))

# ── Early half (vegetative): 1 Oct – 15 May ───────────────────
# Vernalisation, winter dormancy, tillering, green-up
weather_early <- weather_tagged |>
  filter(
    (cal_month %in% c(10, 11, 12)) |     # Oct–Dec
    (cal_month %in% c(1, 2, 3, 4)) |     # Jan–Apr
    (cal_month == 5 & cal_day <= 15)     # 1–15 May
  ) |>
  rename(year = harvest_yr) |>
  agg_weather_slice() |>
  rename_with(~ paste0(.x, "_early"), -c(region, year))

# ── Late half (reproductive): 16 May – 31 Jul ─────────────────
# Heading, anthesis, grain filling
weather_late <- weather_tagged |>
  filter(
    (cal_month == 5 & cal_day >= 16) |   # 16–31 May
    (cal_month %in% c(6, 7))             # Jun–Jul
  ) |>
  rename(year = harvest_yr) |>
  agg_weather_slice() |>
  rename_with(~ paste0(.x, "_late"), -c(region, year))

# ── Diagnostics ───────────────────────────────────────────────
cat("Growing-season obs (region-years): ", nrow(weather_gs),    "\n")
cat("Early-half obs (region-years):     ", nrow(weather_early), "\n")
cat("Late-half obs (region-years):      ", nrow(weather_late),  "\n")

# Quick sanity check: early + late should NOT double-count
# (they are disjoint by construction: May split at 15th)
```

------------------------------------------------------------------------

## 3.3 Model panel assembly

We build `panel_model` by extending `panel_full` (constructed in the previous section) with seasonal weather aggregates, sowing area from the already-loaded `sown` object, soil variables joined via the extended lookup, and the war/occupation dummy.

```{r model-panel-assembly}
library(fixest)
library(modelsummary)

# ── War / occupation dummy ────────────────────────────────────
occupied_oblasts_cyr <- c("Донецька", "Луганська",
                           "Запорізька", "Херсонська")

# ── Build panel_model explicitly from components ──────────────
panel_model <- yield |>
  # Harvest ratio (from harv and sown)
  left_join(harv |> select(region, year, harv_ha), by = c("region", "year")) |>
  left_join(sown |> select(region, year, sown_ha), by = c("region", "year")) |>
  mutate(harv_ratio = pmin(harv_ha / sown_ha, 1.05)) |>
  # Region lookup (Cyrillic → Latin)
  left_join(region_lookup, by = c("region" = "region_cyr")) |>
  # Weather (full year)
  left_join(weather_annual_full, by = c("region_lat" = "region", "year")) |>
  # Seasonal weather
  left_join(weather_gs,    by = c("region_lat" = "region", "year")) |>
  left_join(weather_early, by = c("region_lat" = "region", "year")) |>
  left_join(weather_late,  by = c("region_lat" = "region", "year")) |>
  # Soil
  left_join(soil_joined, by = c("region" = "region_cyr")) |>
  # Derived variables
  mutate(
    war_east   = as.integer(region %in% occupied_oblasts_cyr & year >= 2022),
    log_yield  = log(yield_cpha),
    log_sown   = log(sown_ha + 1),
    region_id  = as.factor(region),
    year_id    = as.factor(year)
  ) |>
  filter(!is.na(yield_cpha), !is.na(days_hot_crit))

# ── Diagnostics ───────────────────────────────────────────────
cat("Model panel rows: ", nrow(panel_model),            "\n")
cat("Regions:          ", n_distinct(panel_model$region_id), "\n")
cat("Years:            ", n_distinct(panel_model$year_id),   "\n")
cat("War-dummy obs:    ", sum(panel_model$war_east),         "\n")
cat("Missing sown_ha:  ", sum(is.na(panel_model$sown_ha)),   "\n")
```

------------------------------------------------------------------------

## 3.4 Model 1 — Full kitchen-sink model

The first model includes all available predictors: full-year weather aggregates (carried over from `panel_full`), sowing area, and the war/occupation dummy. Region and year fixed effects absorb all time-invariant regional heterogeneity (soil quality, geography, historical crop technology) and all common annual shocks (national policy, global climate trends). Because soil variables do not vary within a region over time they are perfectly collinear with region FE and therefore not entered explicitly in any model.

```{r model1}
m1 <- feols(
  yield_cpha ~
    days_hot_crit + days_cold_crit +
    sum_heat_noncrit + sum_cold_noncrit +
    total_precip + total_et0 +
    mean_wind + water_deficit +
    log_sown + war_east
  | region_id + year_id,
  data    = panel_model,
  cluster = ~region_id
)

summary(m1)
```

### 3.4.1 Joint significance test (consistency of all weather variables)

We test whether the complete set of weather regressors is jointly significant after accounting for fixed effects. This is a direct test of Hypothesis 1: if the F-statistic is large and the p-value is small we reject the null that meteorological factors have no collective effect on yield.

```{r model1-ftest}
# Joint Wald F-test: all eight weather variables simultaneously.
# H0: beta_hot = beta_cold = beta_heat = beta_cold_nc =
#     beta_precip = beta_et0 = beta_wind = beta_wdef = 0
wald(m1,
     keep = c("days_hot_crit", "days_cold_crit",
              "sum_heat_noncrit", "sum_cold_noncrit",
              "total_precip", "total_et0",
              "mean_wind", "water_deficit"),
     vcov = "cluster")
```

------------------------------------------------------------------------

## 3.5 Model 1-log — Log-transformed yield (robustness check)

Re-running Model 1 with `log(yield_cpha)` as the dependent variable checks whether results are sensitive to the functional form. Log specification reduces the leverage of high-yield outlier years and allows coefficients to be interpreted as approximate percentage effects.

```{r model1-log}
m1_log <- feols(
  log_yield ~
    days_hot_crit + days_cold_crit +
    sum_heat_noncrit + sum_cold_noncrit +
    total_precip + total_et0 +
    mean_wind + water_deficit +
    log_sown + war_east
  | region_id + year_id,
  data    = panel_model,
  cluster = ~region_id
)

summary(m1_log)
```

------------------------------------------------------------------------

## 3.6 Model 2 — Without sowing area; full-year vs growing-season weather

Removing sowing area (Model 2a) tests whether it contributed independent information or was merely a noisy proxy for regional scale already captured by region FE. We then replace the full-year weather aggregates with growing-season-only aggregates (Model 2b, window: 10 Sep – 31 Jul). If Model 2b fits better than 2a, growing- season weather is more informative than the full calendar year — supporting Hypothesis 2 (winter weather matters).

```{r model2}
# 2a: full-year weather, sowing area removed
m2a <- feols(
  yield_cpha ~
    days_hot_crit + days_cold_crit +
    sum_heat_noncrit + sum_cold_noncrit +
    total_precip + total_et0 +
    mean_wind + water_deficit +
    war_east
  | region_id + year_id,
  data    = panel_model,
  cluster = ~region_id
)

# 2b: growing-season weather (Sep–Jul), no sowing area
m2b <- feols(
  yield_cpha ~
    days_hot_crit_gs + days_cold_crit_gs +
    sum_heat_noncrit_gs + sum_cold_noncrit_gs +
    total_precip_gs + total_et0_gs +
    mean_wind_gs + water_deficit_gs +
    war_east
  | region_id + year_id,
  data    = panel_model,
  cluster = ~region_id
)

# ── Fit comparison table ──────────────────────────────────────
fit_m1_m2 <- tibble(
  Model        = c("M1  (full-yr + sown)",
                   "M2a (full-yr, no sown)",
                   "M2b (growing season)"),
  N            = c(nobs(m1),  nobs(m2a),  nobs(m2b)),
  AIC          = c(AIC(m1),   AIC(m2a),   AIC(m2b)),
  BIC          = c(BIC(m1),   BIC(m2a),   BIC(m2b)),
  R2_overall   = c(r2(m1)[["r2"]],   r2(m2a)[["r2"]],   r2(m2b)[["r2"]]),
  R2_within    = c(r2(m1)[["wr2"]],  r2(m2a)[["wr2"]],  r2(m2b)[["wr2"]]),
  R2_adj       = c(r2(m1)[["ar2"]],  r2(m2a)[["ar2"]],  r2(m2b)[["ar2"]])
)

knitr::kable(fit_m1_m2, digits = 3,
  caption = "Model fit: full-year weather vs growing-season weather")
```

Lower AIC/BIC and higher within-R² for M2b would support Hypothesis 2: weather during the growing season (including the winter dormancy period) is more predictive than the full calendar year, implying that winter temperature shocks are a meaningful driver of final yield.

------------------------------------------------------------------------

## 3.7 Model 3 — Split growing season (early vs late half)

Model 3 replaces each growing-season aggregate with two separate parameters — one for the early half (Oct – 15 May: vernalisation, dormancy, tillering) and one for the late half (16 May – 31 Jul: heading, anthesis, grain filling). This directly tests Hypothesis 3: if early-half and late-half coefficients differ significantly, the *timing* of weather stress matters and simple annual totals discard useful information.

```{r model3}
m3 <- feols(
  yield_cpha ~
    # Early half (Oct – 15 May)
    days_hot_crit_early + days_cold_crit_early +
    sum_heat_noncrit_early + sum_cold_noncrit_early +
    total_precip_early + total_et0_early +
    mean_wind_early + water_deficit_early +
    # Late half (16 May – 31 Jul)
    days_hot_crit_late + days_cold_crit_late +
    sum_heat_noncrit_late + sum_cold_noncrit_late +
    total_precip_late + total_et0_late +
    mean_wind_late + water_deficit_late +
    war_east
  | region_id + year_id,
  data    = panel_model,
  cluster = ~region_id
)

summary(m3)
```

### 3.7.1 Fit comparison: M2b vs M3

```{r model3-fit}
fit_m2b_m3 <- tibble(
  Model      = c("M2b (growing season, pooled)",
                 "M3  (split: early + late)"),
  N          = c(nobs(m2b),  nobs(m3)),
  AIC        = c(AIC(m2b),   AIC(m3)),
  BIC        = c(BIC(m2b),   BIC(m3)),
  R2_within  = c(r2(m2b)[["wr2"]], r2(m3)[["wr2"]]),
  R2_adj     = c(r2(m2b)[["ar2"]], r2(m3)[["ar2"]])
)

knitr::kable(fit_m2b_m3, digits = 3,
  caption = "Model fit: pooled growing season vs split season")
```

### 3.7.2 Wald test — do early and late half coefficients differ?

The joint equality test below asks whether each weather variable has the same effect in the early half as in the late half. Rejection of H₀ provides formal evidence that phase-specific aggregation adds predictive content beyond the pooled growing-season model.

```{r model3-wald}
# H0: for each variable, beta_early == beta_late
wald(m3,
     c("days_hot_crit_early   = days_hot_crit_late",
       "days_cold_crit_early  = days_cold_crit_late",
       "sum_heat_noncrit_early = sum_heat_noncrit_late",
       "sum_cold_noncrit_early = sum_cold_noncrit_late",
       "total_precip_early    = total_precip_late",
       "total_et0_early       = total_et0_late",
       "mean_wind_early       = mean_wind_late",
       "water_deficit_early   = water_deficit_late"),
     vcov = "cluster")
```

------------------------------------------------------------------------

## 3.8 Full model comparison table

The table below places all five specifications side by side. Stars denote statistical significance with clustered standard errors at the oblast level. The bottom rows report the within-R² (variation explained after demeaning by region and year) — the most relevant fit metric for two-way FE models.

```{r model-table, message=FALSE}
coef_map <- c(
  # Full-year (M1, M1-log, M2a)
  days_hot_crit        = "Hot critical days (full yr)",
  days_cold_crit       = "Cold critical days (full yr)",
  sum_heat_noncrit     = "Accum. heat (full yr)",
  sum_cold_noncrit     = "Accum. cold (full yr)",
  total_precip         = "Precipitation (full yr)",
  total_et0            = "ET0 (full yr)",
  mean_wind            = "Wind gusts (full yr)",
  water_deficit        = "Water deficit (full yr)",
  log_sown             = "log(Sown area)",
  # Growing season (M2b)
  days_hot_crit_gs     = "Hot critical days (GS)",
  days_cold_crit_gs    = "Cold critical days (GS)",
  sum_heat_noncrit_gs  = "Accum. heat (GS)",
  sum_cold_noncrit_gs  = "Accum. cold (GS)",
  total_precip_gs      = "Precipitation (GS)",
  total_et0_gs         = "ET0 (GS)",
  mean_wind_gs         = "Wind (GS)",
  water_deficit_gs     = "Water deficit (GS)",
  # Early half (M3)
  days_hot_crit_early   = "Hot critical days (Oct–May)",
  days_cold_crit_early  = "Cold critical days (Oct–May)",
  sum_heat_noncrit_early = "Accum. heat (Oct–May)",
  sum_cold_noncrit_early = "Accum. cold (Oct–May)",
  total_precip_early    = "Precipitation (Oct–May)",
  total_et0_early       = "ET0 (Oct–May)",
  mean_wind_early       = "Wind (Oct–May)",
  water_deficit_early   = "Water deficit (Oct–May)",
  # Late half (M3)
  days_hot_crit_late    = "Hot critical days (May–Jul)",
  days_cold_crit_late   = "Cold critical days (May–Jul)",
  sum_heat_noncrit_late = "Accum. heat (May–Jul)",
  sum_cold_noncrit_late = "Accum. cold (May–Jul)",
  total_precip_late     = "Precipitation (May–Jul)",
  total_et0_late        = "ET0 (May–Jul)",
  mean_wind_late        = "Wind (May–Jul)",
  water_deficit_late    = "Water deficit (May–Jul)",
  # Common
  war_east             = "War × East oblasts (2022+)"
)

modelsummary(
  list(
    "M1\nfull-yr+sown" = m1,
    "M1-log\n(log Y)"  = m1_log,
    "M2a\nfull-yr"     = m2a,
    "M2b\ngrow.season" = m2b,
    "M3\nsplit season"  = m3
  ),
  coef_map  = coef_map,
  stars     = c("*" = 0.10, "**" = 0.05, "***" = 0.01),
  gof_map   = c("nobs", "r.squared", "adj.r.squared", "AIC", "BIC"),
  title     = "Panel fixed-effects models — winter wheat yield, Ukraine 2015–2024",
  notes     = paste(
    "Two-way FE: region + year. Standard errors clustered at oblast level.",
    "M1-log: dependent variable is log(yield_cpha).",
    "GS = growing season (10 Sep – 31 Jul of harvest year).",
    "Early half: 1 Oct – 15 May. Late half: 16 May – 31 Jul.",
    "*** p<0.01, ** p<0.05, * p<0.10."
  )
)
```

------------------------------------------------------------------------

# Methods.

Explain accurately what methods you are using to solve the tasks of interest.

To solve these tasks, we apply the following: 1. **Pooled OLS & Polynomial Regression:** To capture the parabolic effects of temperature and radiation. 2. **Segmented (Piecewise) Regression:** Specifically for our "Extra Days" hypothesis to model heat stress above $30^{\circ}C$. 3. **Fixed Effects (FE) Model:** To control for time-invariant regional factors like soil quality across Ukraine. 4. **Moderation Analysis:** Using interaction terms ($Yield \times Factor$) to see how natural factors alter the sensitivity of price to yield]. 5. **Correlation Matrix:** To test the $H_1$ of Hypothesis 4 and decide whether to omit cloudiness in favor of rain/shine to maintain model parsimony.

------------------------------------------------------------------------

# Results.

Describe the results you already have obtained at this stage.

At this stage (April 2, 2026), we have: - Completed the collection of meteorological data via Open-Meteo API and price data from Minfin. - Constructed a unified panel dataset aligned by weekly averages. - Initiated the classification of "Hot" and "Cold" days to test our Extreme Day hypothesis.

------------------------------------------------------------------------

# Conclusions.

What conclusions have you already drawn?

Early findings indicate that weather impacts are indeed non-linear, supporting our use of quadratic forms in the model. We have also concluded that high-frequency daily data is necessary to accurately count "extreme" events before aggregating them into the final econometric model.
