---
title: "Agrobusiness"
author: "Kate, Anastasiia, Max"
date: "2026-04-02"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
header-includes:
  - \usepackage{fontspec}
  - \setmainfont{DejaVu Sans}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 7, dpi = 150)
```

------------------------------------------------------------------------

# Hypotheses and objectives description

This project investigates the drivers of winter wheat yield volatility across Ukrainian oblasts using panel data for 2015–2024. The core research question is: which meteorological factors actually determine yield outcomes, and which are redundant for practical data collection purposes? The goal is to provide agribusinesses with a parsimonious, empirically grounded framework — identifying what weather data is mandatory to collect and what can be safely dropped.

### Research Hypotheses

**1. Traditional Significance and Input Factors:**

$H_0$: Meteorological factors have no statistically significant effect on wheat yield.

$H_1$: Key weather variables — particularly extreme temperature events and precipitation — have significant effects on yield, with the relationship following a non-linear pattern where both heat and cold extremes are detrimental.

**2. Seasonal Relevancy (The "Window" Question):**

$H_0$: Weather aggregated over the full calendar year is no less predictive than weather aggregated over the growing season only.

$H_1$: The growing season window (September–July) captures the agronomically relevant stress periods more precisely than full-year averages, yielding better model fit.

**3. Data Granularity and Extreme Day Classification:**

$H_0$: Counting extreme temperature days does not provide more predictive power than simple mean temperature.

$H_1$: The frequency of extreme events — days above the critical heat threshold and days below the critical cold threshold — is a more accurate predictor of yield shocks than mean temperature alone, justifying the collection of daily rather than weekly data.

**4. Redundancy and Variable Selection:**

$H_0$: All weather variables (radiation, ET₀, dewpoint, temperature, precipitation) contribute independent predictive information.

$H_1$: Solar radiation, ET₀, and dewpoint are strongly collinear with temperature metrics and precipitation, making them redundant. A parsimonious model retaining only temperature stress metrics, precipitation, wind, and soil moisture performs equivalently.

**5. Soil Quality as a Moderator:**

$H_0$: Soil properties have no effect on yield once regional fixed effects are controlled for.

$H_1$: Static soil properties — particularly organic matter and sand content — affect average yield levels and moderate the sensitivity of yield to weather shocks.

------------------------------------------------------------------------

# Data Description and Exploratory Analysis

## 1. Introduction

This study investigates the impact of weather factors on winter wheat yield in Ukraine using panel data for 2015–2024 at the oblast level. The dependent variable is yield in centners per hectare. Key explanatory variables are derived from daily temperature and soil moisture data following a theoretically motivated aggregation scheme. Soil properties and area variables serve as controls.

------------------------------------------------------------------------

## 2. Data Description

This section describes all data sources used in the study: crop statistics (annual, oblast level), daily weather records, static soil properties, and derived weather metrics constructed following the research hypothesis.

### 2.1 Crop Data (annual, by oblast)

| File | Variable | Units | Explanation |
|------------------|------------------|------------------|------------------|
| `crop_capacity.csv` | Yield | centners/ha | Dependent variable. Already calculated by statistical authorities. |
| `harvested_area.csv` | Harvested area | hectares | Area from which crop was actually harvested. |
| `sowing_area.csv` | Sown area | hectares | Area planted (may differ from harvested due to losses). |

```{r}
library(tidyverse)
library(lubridate)
library(patchwork)

data_path <- "."
years     <- as.character(2015:2025)

META_COLS <- c("indicator", "region", "unit", "scale",
               "decimals", "special", "note_ts", "note_obs")

read_panel <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names      = c(META_COLS, years),
    skip           = 1,
    show_col_types = FALSE
  ) |>
    dplyr::select(region, all_of(years)) |>
    pivot_longer(cols = all_of(years), names_to = "year",
                 values_to = value_name) |>
    mutate(
      year         = as.integer(year),
      !!value_name := suppressWarnings(as.numeric(.data[[value_name]]))
    ) |>
    filter(region != "Україна")
}

read_ukraine <- function(file, value_name) {
  read_csv(
    file.path(data_path, file),
    col_names      = c(META_COLS, years),
    skip           = 1,
    show_col_types = FALSE
  ) |>
    filter(region == "Україна") |>
    dplyr::select(region, all_of(years)) |>
    pivot_longer(cols = all_of(years), names_to = "year",
                 values_to = value_name) |>
    mutate(year = as.integer(year),
           !!value_name := suppressWarnings(as.numeric(.data[[value_name]])))
}

yield     <- read_panel("crop_capacity.csv", "yield_cpha")
harv      <- read_panel("harvested_area.csv", "harv_kha") |>
               mutate(harv_ha = harv_kha * 1000) |> 
               dplyr::select(-harv_kha)
sown      <- read_panel("sowing_area.csv", "sown_kha") |>
               mutate(sown_ha = sown_kha * 1000) |> 
               dplyr::select(-sown_kha)

ukr_yield <- read_ukraine("crop_capacity.csv", "yield_cpha")

region_groups <- tribble(
  ~region,              ~group,
  "Волинська",          "West",
  "Закарпатська",       "West",
  "Івано-Франківська",  "West",
  "Львівська",          "West",
  "Рівненська",         "West",
  "Тернопільська",      "West",
  "Хмельницька",        "West",
  "Чернівецька",        "West",
  "Вінницька",          "Centre",
  "Дніпропетровська",   "Centre",
  "Житомирська",        "Centre",
  "Київська",           "Centre",
  "Кіровоградська",     "Centre",
  "Полтавська",         "Centre",
  "Сумська",            "Centre",
  "Харківська",         "Centre",
  "Черкаська",          "Centre",
  "Чернігівська",       "Centre",
  "Донецька",           "East",
  "Запорізька",         "East",
  "Луганська",          "East",
  "Миколаївська",       "East",
  "Одеська",            "East",
  "Херсонська",         "East"
)

GROUP_COLORS <- c(West = "#2c7bb6", Centre = "#d7191c", East = "#1a9641")

theme_small_multiple <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      panel.border      = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.line.x       = element_line(colour = "grey70", linewidth = 0.4),
      axis.ticks.x      = element_line(colour = "grey70", linewidth = 0.3),
      axis.ticks.y      = element_blank(),
      axis.text.x       = element_text(size = 7, colour = "grey40",
                                       margin = margin(t = 2)),
      axis.text.y       = element_text(size = 7, colour = "grey40"),
      axis.title        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8,
                                       colour = "grey20", margin = margin(b = 3)),
      strip.background  = element_blank(),
      strip.placement   = "outside",
      legend.position   = "none",
      plot.title        = element_text(face = "bold", size = 11,
                                       colour = "grey10", margin = margin(b = 4)),
      plot.subtitle     = element_text(size = 8.5, colour = "grey40",
                                       margin = margin(b = 8)),
      plot.caption      = element_text(size = 7, colour = "grey55",
                                       hjust = 0, margin = margin(t = 6)),
      plot.margin       = margin(10, 12, 8, 10)
    )
}

plot_small_multiple <- function(df, group_map, value_col,
                                group    = NULL,
                                title    = "",
                                subtitle = "",
                                caption  = "") {

  df <- df |>
    left_join(group_map, by = "region") |>
    rename(value = !!sym(value_col))

  focal_regions <- if (!is.null(group)) {
    group_map |> filter(group == !!group) |> pull(region)
  } else {
    unique(df$region)
  }

  focal_color <- if (!is.null(group)) GROUP_COLORS[[group]] else "grey30"

  bg_data <- tibble(facet = focal_regions) |>
    cross_join(
      df |> filter(!is.na(value)) |> dplyr::select(region, year, value)
    )

  focal_data <- df |>
    filter(region %in% focal_regions, !is.na(value)) |>
    mutate(facet = region)

  ggplot() +
    geom_line(
      data    = bg_data |> filter(region != facet),
      mapping = aes(x = year, y = value, group = region),
      colour  = "grey85", linewidth = 0.35, alpha = 0.9
    ) +
    geom_line(
      data    = focal_data,
      mapping = aes(x = year, y = value, group = facet),
      colour  = focal_color, linewidth = 1.1, lineend = "round"
    ) +
    facet_wrap(~facet, scales = "free_y") +
    scale_x_continuous(
      breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
      labels = c("'15", "'17", "'19", "'21", "'23", "'25")
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.10))) +
    labs(title = title, subtitle = subtitle, caption = caption) +
    theme_small_multiple()
}


p_yield_west <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "West",
  title    = "Winter wheat yield — Western oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_yield_centre <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "Centre",
  title    = "Winter wheat yield — Central oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_yield_east <- plot_small_multiple(
  yield, region_groups, "yield_cpha", "East",
  title    = "Winter wheat yield — Eastern oblasts",
  subtitle = "Centners per hectare, 2015–2025. Grey lines = all other regions.",
  caption  = "Source: State Statistics Service of Ukraine"
)


loss_rate <- yield |>
  left_join(sown |> dplyr::select(region, year, sown_ha), by = c("region", "year")) |>
  left_join(harv |> dplyr::select(region, year, harv_ha), by = c("region", "year")) |>
  mutate(
    loss_rate = (sown_ha - harv_ha) / sown_ha,
    loss_rate = if_else(loss_rate < 0, 0, loss_rate),
    loss_rate = if_else(loss_rate > 0.5, 0.5, loss_rate)
  ) |>
  filter(!is.na(loss_rate))

p_loss_west <- plot_small_multiple(
  loss_rate, region_groups, "loss_rate", "West",
  title    = "Crop loss rate — Western oblasts",
  subtitle = "(Sown area − Harvested area) / Sown area. 0% = perfect harvest.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_loss_centre <- plot_small_multiple(
  loss_rate, region_groups, "loss_rate", "Centre",
  title    = "Crop loss rate — Central oblasts",
  subtitle = "(Sown area − Harvested area) / Sown area. 0% = perfect harvest.",
  caption  = "Source: State Statistics Service of Ukraine"
)

p_loss_east <- plot_small_multiple(
  loss_rate, region_groups, "loss_rate", "East",
  title    = "Crop loss rate — Eastern oblasts",
  subtitle = "(Sown area − Harvested area) / Sown area. 0% = perfect harvest.",
  caption  = "Source: State Statistics Service of Ukraine"
)


oblast_area_km2 <- tribble(
  ~region,              ~total_km2,
  "Вінницька",          26513,
  "Волинська",          20144,
  "Дніпропетровська",   31914,
  "Донецька",           26517,
  "Житомирська",        29832,
  "Закарпатська",       12777,
  "Запорізька",         27180,
  "Івано-Франківська",  13900,
  "Київська",           28131,
  "Кіровоградська",     24588,
  "Луганська",          26683,
  "Львівська",          21833,
  "Миколаївська",       24598,
  "Одеська",            33310,
  "Полтавська",         28748,
  "Рівненська",         20047,
  "Сумська",            23832,
  "Тернопільська",      13823,
  "Харківська",         31415,
  "Херсонська",         28461,
  "Хмельницька",        20645,
  "Черкаська",          20900,
  "Чернівецька",         8097,
  "Чернігівська",       31865
) |>
  mutate(total_ha = total_km2 * 100)

sown_pct <- sown |>
  left_join(oblast_area_km2 |> dplyr::select(region, total_ha), by = "region") |>
  mutate(sown_pct = sown_ha / total_ha * 100) |>
  filter(!is.na(sown_pct))

p_sownpct_west <- plot_small_multiple(
  sown_pct, region_groups, "sown_pct", "West",
  title    = "Sown wheat area as % of total oblast area — West",
  subtitle = "% of total oblast land area planted with winter wheat, 2015–2025.",
  caption  = "Oblast areas: official administrative data. Sown area: SSU."
)

p_sownpct_centre <- plot_small_multiple(
  sown_pct, region_groups, "sown_pct", "Centre",
  title    = "Sown wheat area as % of total oblast area — Centre",
  subtitle = "% of total oblast land area planted with winter wheat, 2015–2025.",
  caption  = "Oblast areas: official administrative data. Sown area: SSU."
)

p_sownpct_east <- plot_small_multiple(
  sown_pct, region_groups, "sown_pct", "East",
  title    = "Sown wheat area as % of total oblast area — East",
  subtitle = "% of total oblast land area planted with winter wheat, 2015–2025.",
  caption  = "Oblast areas: official administrative data. Sown area: SSU."
)

sown_pct_avg <- sown_pct |>
  left_join(region_groups, by = "region") |>
  group_by(region, group) |>
  summarise(mean_pct = mean(sown_pct, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_pct) |>
  mutate(region = factor(region, levels = region))

p_sownpct_bar <- ggplot(sown_pct_avg,
                        aes(x = mean_pct, y = region, fill = group)) +
  geom_col(width = 0.72, alpha = 0.88) +
  geom_text(aes(label = paste0(round(mean_pct, 1), "%")),
            hjust = -0.1, size = 2.6, colour = "grey30") +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.12)),
    labels = scales::label_number(suffix = "%")
  ) +
  labs(
    title    = "Average sown wheat area as % of total oblast area (2015–2025)",
    subtitle = "Oblasts ordered by intensity. Colour = geographical group.",
    caption  = "Oblast areas: official administrative data. Sown area: SSU.",
    fill     = "Group"
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    axis.text.y        = element_text(size = 8),
    legend.position    = "right",
    legend.text        = element_text(size = 8),
    legend.title       = element_text(size = 8, face = "bold")
  )


rep_oblasts <- yield |>
  left_join(region_groups, by = "region") |>
  filter(!is.na(group), !is.na(yield_cpha)) |>
  group_by(group, region) |>
  summarise(mean_y = mean(yield_cpha, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_y, n = 1, by = group)

rep_data <- yield |>
  inner_join(rep_oblasts |> dplyr::select(group, region), by = "region") |>
  filter(!is.na(yield_cpha))

p_representative <- ggplot() +
  geom_line(
    data    = ukr_yield |> filter(!is.na(yield_cpha)),
    mapping = aes(x = year, y = yield_cpha),
    colour  = "grey65", linewidth = 0.7, linetype = "dashed"
  ) +
  geom_line(
    data    = rep_data,
    mapping = aes(x = year, y = yield_cpha, colour = group),
    linewidth = 1.2, lineend = "round"
  ) +
  facet_wrap(~group) +
  scale_colour_manual(values = GROUP_COLORS) +
  scale_x_continuous(
    breaks = c(2015, 2017, 2019, 2021, 2023, 2025),
    labels = c("'15", "'17", "'19", "'21", "'23", "'25")
  ) +
  labs(
    title    = "Most productive representative oblast per geographical group",
    subtitle = "Dashed grey = Ukraine national average. Coloured = representative oblast (highest mean yield).",
    caption  = "Source: State Statistics Service of Ukraine"
  ) +
  theme_small_multiple(base_size = 10) +
  theme(strip.text = element_text(face = "bold", size = 10, colour = "grey15"))


print(p_yield_west)
print(p_yield_centre)
print(p_yield_east)

print(p_loss_west)
print(p_loss_centre)
print(p_loss_east)

print(p_sownpct_west)
print(p_sownpct_centre)
print(p_sownpct_east)
print(p_sownpct_bar)

print(p_representative)
```

#### Yield Patterns and Regional Heterogeneity

The small-multiple yield plots reveal substantial variation across oblasts both in levels and in temporal dynamics. Western regions (Vinnytsia, Khmelnytskyi, Ternopil) consistently show higher yields (45–75 ц/га) with relatively stable trajectories, while eastern oblasts (Odesa, Zaporizhzhia, Donetsk) exhibit lower and more volatile yields (25–55 ц/га), likely reflecting differences in soil quality, moisture availability, and war-related disruptions after 2022.

Critically, the plots demonstrate that **different regions follow different yield trends**. Some oblasts show gradual increases over 2015–2021 (suggesting technology adoption or management improvements), others remain flat, and a few (particularly in the east) display declining pre-war trends. This heterogeneity has direct implications for our modelling strategy: a single common time trend will be insufficient. We will therefore include both a common trend (capturing Ukraine-wide technological progress) and region-specific trends (allowing each oblast's trajectory to diverge).

#### Loss Rates and the Lagged Effect Argument

The loss rate plots — defined as `(sown area − harvested area) / sown area` — show that even in normal years, 2–8% of sown wheat area is lost before harvest. In extreme years (e.g., 2020 in Odesa, 2022 in eastern oblasts), losses exceed 15–20%.

Why does this matter for modelling current-year yield? **Losses in year t−1 can affect yield in year t through farmer behavioural responses.** After a year of high losses, farmers may:

1.  **Increase input intensity** (more fertiliser, better seeds, more frequent fungicide applications) to compensate for lost income or to secure the next harvest
2.  **Alter sowing decisions** (switch to more resilient varieties, adjust planting dates, change crop rotation)
3.  **Face financial constraints** that limit their ability to invest in the following season

Conversely, low-loss years may create complacency or reduce the perceived need for intensive management. These dynamics imply that the loss rate from the previous year — which we denote as `loss_lag` — is a legitimate predictor of current yield, capturing unobserved farmer responses that are not directly measured in our data. We will test this hypothesis explicitly in the modelling section by including `loss_lag` as a regressor.

------------------------------------------------------------------------

## 2.2 Daily Weather Data (`daily_weather.csv`) and Soil Moisture

| Variable | Description | Unit |
|------------------------|------------------------|------------------------|
| `region` | Oblast name | text |
| `date` | Calendar date | YYYY-MM-DD |
| `temperature_2m_max` | Daily maximum temperature | °C |
| `temperature_2m_min` | Daily minimum temperature | °C |
| `temperature_2m_mean` | Daily mean temperature | °C |
| `wind_speed_10m_max` | Daily maximum wind speed at 10m height | m/s |
| `wind_gusts_10m_max` | Daily maximum wind gusts at 10m height | m/s |
| `et0_fao_evapotranspiration` | Reference evapotranspiration (FAO Penman-Monteith) | mm/day |
| `dewpoint_2m_mean` | Daily mean dewpoint temperature | °C |
| `precipitation_sum` | Daily total precipitation | mm |
| `shortwave_radiation_sum` | Daily incoming shortwave solar radiation | MJ/m² |

**Additional data — soil moisture (`soil_moisture.csv`):**

| Variable | Description | Unit |
|------------------------|------------------------|------------------------|
| `region` | Oblast name | text |
| `date` | Calendar date | YYYY-MM-DD |
| `soil_moisture_0_7cm_pct` | Volumetric soil moisture (0–7 cm depth) | \% |

**Notes:** ET₀ represents atmospheric evaporative demand — higher values indicate faster water loss from crops. Dewpoint is a direct measure of air moisture (higher values promote fungal diseases). Data cover 2013–2024 for all 24 oblasts. Soil moisture data are sourced from Open-Meteo ERA5-Land archive.

#### Expected Correlations Among Weather Parameters

Understanding correlations helps us avoid multicollinearity in later models:

-   **ET₀, solar radiation, and temperature** — strongly positively correlated. We will select only ET₀ as the representative.
-   **Precipitation and soil moisture** — moderately correlated (1–3 day lag), but soil moisture captures drought persistence uniquely.
-   **Dewpoint and temperature** — context-dependent correlation; dewpoint carries disease-pressure information.
-   **Wind speed** — largely uncorrelated with others; kept for lodging risk.
-   **Soil moisture vs. temperature/ET₀** — negative correlation (hot conditions deplete soil water).

#### Constructed Weather Metrics (following the research hypothesis)

**Why aggregate daily data?** Our hypothesis posits that **extreme days matter more than averages**. Using daily data directly would create thousands of predictors. The solution is to aggregate into theoretically grounded stress metrics, preserving non-linear "extreme shock" logic while collapsing 365 days into few annual numbers.

**Information loss recovery:** Aggregation loses timing. We recover this by splitting the year into meaningful periods (winter, early spring, late spring) and calculating metrics separately for each window.

For each region and year, **temperature** is transformed as:

$$ t_{dev} = \frac{T_{max} + T_{min}}{2} - 5 $$

where 5°C is the assumed base temperature for winter wheat development.

From this deviation, four temperature metrics are calculated:

| Metric | Definition |
|------------------------------------|------------------------------------|
| `days_hot_critical` | Number of days with $t_{dev} > 20$ (extreme heat stress) |
| `days_cold_critical` | Number of days with $t_{dev} < -15$ (extreme cold stress) |
| `sum_heat_noncrit` | Sum of $t_{dev}$ for days with $0 < t_{dev} \leq 20$ |
| `sum_cold_noncrit` | Sum of $t_{dev}$ for days with $-15 \leq t_{dev} < 0$ |

For **soil moisture**, we apply the **exact same logic** — counting extreme events rather than using raw values:

| Metric | Definition |
|------------------------------------|------------------------------------|
| `days_drought` | Number of days with soil moisture \< 20% (severe water stress) |
| `days_waterlog` | Number of days with soil moisture \> 85% (root asphyxiation risk) |
| `mean_soil_moisture` | Annual average soil moisture (baseline condition) |

**Why these thresholds?** For winter wheat: - **\<20% soil moisture** — permanent wilting point, stomata close, photosynthesis stops - **\>85% soil moisture** — oxygen depletion in root zone, root rot and fungal diseases

This mirrors our temperature logic: just as we count `days_hot_crit` and `days_cold_crit`, we now count `days_drought` and `days_waterlog` as the water-side equivalents.

**Note on trends:** All weather metrics exhibit trends over time (warming climate). These will be absorbed by common time trends (`t`) and region-specific trends (`region × t`) in the modelling section.

```{r fig.width=14, fig.height=9}
# ── Load and aggregate weather data ─────────────────────────
weather <- read_csv("daily_weather.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date), year = year(date)) |>
  filter(year >= 2013, year <= 2024)

# ── Load and aggregate soil moisture data ───────────────────
# Same logic as temperature: count extreme days, not just averages
soil_moisture <- read_csv("soil_moisture.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date), year = year(date)) |>
  filter(year >= 2013, year <= 2024)

soil_moisture_agg <- soil_moisture |>
  mutate(
    # Same extreme-day counting logic as temperature
    drought_day = soil_moisture_0_7cm_pct < 20,
    waterlog_day = soil_moisture_0_7cm_pct > 85
  ) |>
  group_by(region, year) |>
  summarise(
    mean_soil_moisture_pct = mean(soil_moisture_0_7cm_pct, na.rm = TRUE),
    days_drought = sum(drought_day, na.rm = TRUE),
    days_waterlog = sum(waterlog_day, na.rm = TRUE),
    .groups = "drop"
  )

BASE_TEMP <- 5

weather_annual <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - BASE_TEMP,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20,  t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0,  t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    # Thermal stress metrics
    days_hot_crit    = sum(days_hot_crit,  na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit, na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,   na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,   na.rm = TRUE),
    # Raw temperature levels
    mean_tmax        = mean(temperature_2m_max,          na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min,          na.rm = TRUE),
    mean_temp        = mean(t_mean,                      na.rm = TRUE),
    # Water and energy
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    total_radiation  = sum(shortwave_radiation_sum,      na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    mean_dewpoint    = mean(dewpoint_2m_mean,            na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(soil_moisture_agg, by = c("region", "year"))

# Order regions W→E by mean temperature
region_order <- weather_annual |>
  group_by(region) |>
  summarise(mt = mean(mean_temp, na.rm = TRUE)) |>
  arrange(mt) |>
  pull(region)

weather_annual <- weather_annual |>
  mutate(region = factor(region, levels = region_order),
         year   = factor(year))

# ── Shared theme ────────────────────────────────────────────
theme_heatmap <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid        = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.text.x       = element_text(size = 7.5, angle = 45, hjust = 1,
                                       colour = "grey35"),
      axis.text.y       = element_text(size = 7,   colour = "grey30"),
      axis.title        = element_blank(),
      axis.ticks        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8.5,
                                       colour = "grey15"),
      strip.background  = element_blank(),
      plot.title        = element_text(face = "bold", size = 10,
                                       colour = "grey10",
                                       margin = margin(b = 2)),
      plot.subtitle     = element_text(size = 7.5, colour = "grey45",
                                       margin = margin(b = 5)),
      legend.text       = element_text(size = 7,   colour = "grey35"),
      legend.title      = element_blank(),
      legend.key.width  = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin       = margin(6, 8, 6, 6)
    )
}

# ── Helper: one heatmap tile ────────────────────────────────
make_tile <- function(df, metric, palette, direction = 1,
                      title, subtitle) {
  ggplot(df, aes(x = year, y = region, fill = .data[[metric]])) +
    geom_tile(colour = "white", linewidth = 0.35) +
    scale_fill_distiller(palette   = palette,
                         direction = direction,
                         guide     = guide_colorbar(
                           barwidth  = 0.45,
                           barheight = 7,
                           ticks     = FALSE)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(title = title, subtitle = subtitle) +
    theme_heatmap()
}

# ── Temperature stress metrics ──────────────────────────────
p1 <- make_tile(weather_annual, "days_hot_crit", "YlOrRd", 1,
  "Days with extreme heat stress",
  "t_dev > 20 °C (mean daily temp > 25 °C) — count of days per year")

p2 <- make_tile(weather_annual, "days_cold_crit", "Blues", -1,
  "Days with extreme cold stress",
  "t_dev < −15 °C (mean daily temp < −10 °C) — count of days per year")

p3 <- make_tile(weather_annual, "sum_heat_noncrit", "YlOrBr", 1,
  "Accumulated non-critical heat (growing degree-days)",
  "Sum of t_dev for days where 0 < t_dev ≤ 20 °C — thermal budget for growth")

p4 <- make_tile(weather_annual, "sum_cold_noncrit", "PuBu", -1,
  "Accumulated non-critical cold (vernalisation budget)",
  "Sum of t_dev for days where −15 ≤ t_dev < 0 °C — chilling requirement")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Temperature stress metrics by region and year",
    subtitle = paste0("Derived from daily Tmax / Tmin. Base temperature = ",
                      BASE_TEMP, " °C. Regions ordered cool → warm (W/N to E/S)."),
    caption  = "Source: daily_weather.csv",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.subtitle   = element_text(size = 9, colour = "grey45"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
# ── Raw temperature levels ──────────────────────────────────
p_tmax <- make_tile(weather_annual, "mean_tmax", "RdYlBu", -1,
  "Mean daily Tmax (°C)",
  "Annual average of daily maximum temperature")

p_tmin <- make_tile(weather_annual, "mean_tmin", "RdYlBu", -1,
  "Mean daily Tmin (°C)",
  "Annual average of daily minimum temperature")

p_tmean <- make_tile(weather_annual, "mean_temp", "RdYlBu", -1,
  "Mean daily temperature (°C)",
  "Annual average of (Tmax + Tmin) / 2")

(p_tmax | p_tmin | p_tmean) +
  plot_annotation(
    title   = "Raw temperature levels by region and year",
    caption = "Source: daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
# ── Precipitation ───────────────────────────────────────────
make_tile(weather_annual, "total_precip", "BrBG", 1,
  "Total annual precipitation (mm)",
  "Sum of daily precipitation. Green = wetter, brown = drier.") +
  plot_annotation(
    caption = "Source: daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
# ── Evapotranspiration ──────────────────────────────────────
make_tile(weather_annual, "total_et0", "YlOrRd", 1,
  "Total annual ET₀ (mm/year)",
  "Sum of daily reference evapotranspiration") +
  plot_annotation(
    title   = "Reference evapotranspiration (ET₀) by region and year",
    caption = "Source: daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
# ── Wind speed ──────────────────────────────────────────────
make_tile(weather_annual, "mean_wind", "PuRd", 1,
  "Mean annual maximum wind gusts (m/s)",
  "Annual average of daily maximum wind gust at 10 m height") +
  plot_annotation(
    caption = "Source: daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

```{r}
# ── Solar radiation ─────────────────────────────────────────
make_tile(weather_annual, "total_radiation", "YlOrRd", 1,
  "Total annual solar radiation (MJ/m²)",
  "Sum of daily shortwave radiation") +
  plot_annotation(
    title   = "Solar radiation by region and year",
    caption = "Source: daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

```{r}
# ── Soil moisture metrics (same logic as temperature) ───────
p_soil_mean <- make_tile(weather_annual, "mean_soil_moisture_pct", "BrBG", 1,
  "Mean annual soil moisture (0-7 cm, %)",
  "Average volumetric soil moisture. Brown = dry, green = wet.")

p_drought <- make_tile(weather_annual, "days_drought", "YlOrRd", 1,
  "Days with drought conditions",
  "Soil moisture < 20% — permanent wilting point (same logic as days_hot_crit)")

p_waterlog <- make_tile(weather_annual, "days_waterlog", "Blues", -1,
  "Days with waterlogging conditions",
  "Soil moisture > 85% — root asphyxiation risk (mirroring days_cold_crit logic)")

(p_soil_mean | p_drought) / (p_waterlog | plot_spacer()) +
  plot_annotation(
    title   = "Soil moisture metrics by region and year",
    subtitle = "Following same extreme-day counting logic as temperature stress metrics",
    caption = "Source: soil_moisture.csv (ERA5-Land via Open-Meteo)",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.subtitle   = element_text(size = 9, colour = "grey45"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

**Note on waterlogging detection:** The soil moisture heatmaps show virtually zero days exceeding the 85% waterlogging threshold across all regions and years. This does not imply that floods or waterlogging events did not occur in Ukraine during 2015–2023. Rather, this reflects two limitations of our data: first, the ERA5-Land reanalysis product smooths extreme moisture events, potentially underestimating short-term saturation episodes; second, the 0–7 cm surface layer drains rapidly and does not capture waterlogging in the deeper root zone (20–60 cm), where crop damage actually occurs. Consequently, our days_waterlog metric should be interpreted as a measure of extreme, persistent surface saturation rather than a comprehensive indicator of flood risk. For practical agronomic applications, precipitation intensity (e.g., daily rainfall \>40 mm) or multi-day cumulative rainfall may serve as more reliable proxies for waterlogging risk.

------------------------------------------------------------------------

```{r}
# ── Summary table ───────────────────────────────────────────
weather_annual |>
  group_by(region) |>
  summarise(
    `Hot days`     = round(mean(days_hot_crit, na.rm = TRUE),    1),
    `Cold days`    = round(mean(days_cold_crit, na.rm = TRUE),   1),
    `Heat sum`     = round(mean(sum_heat_noncrit, na.rm = TRUE), 0),
    `Cold sum`     = round(mean(sum_cold_noncrit, na.rm = TRUE), 0),
    `Precip (mm)`  = round(mean(total_precip, na.rm = TRUE),     0),
    `ET₀ (mm/yr)`  = round(mean(total_et0, na.rm = TRUE),        0),
    `Wind (m/s)`   = round(mean(mean_wind, na.rm = TRUE),        1),
    `Soil moisture (%)` = round(mean(mean_soil_moisture_pct, na.rm = TRUE), 1),
    `Drought days` = round(mean(days_drought, na.rm = TRUE),     1),
    `Waterlog days` = round(mean(days_waterlog, na.rm = TRUE),    1)
  ) |>
  arrange(desc(`Hot days`)) |>
  knitr::kable(caption = "Mean annual weather and soil moisture metrics by oblast (2013–2024, ordered by heat stress days)")
```

------------------------------------------------------------------------

### 2.3 Soil Data (`soil_data.csv`)

Static soil properties by oblast:

| Variable         | Description                        | Unit     |
|------------------|------------------------------------|----------|
| `oblast`         | Oblast name                        | text     |
| `pH`             | Soil acidity                       | unitless |
| `organic_matter` | Organic matter content             | \%       |
| `total_nitrogen` | Total nitrogen content             | \%       |
| `cec`            | Cation Exchange Capacity           | meq/100g |
| `clay`           | Clay content (\<0.002 mm)          | \%       |
| `sand`           | Sand content (0.05–2.0 mm)         | \%       |
| `silt`           | Silt content (0.002–0.05 mm)       | \%       |
| `ec`             | Electrical conductivity (salinity) | dS/m     |
| `esp`            | Exchangeable Sodium Percentage     | \%       |
| `c_n_ratio`      | Carbon to Nitrogen ratio           | unitless |

**Notes:** Optimal pH for winter wheat is 6.0–7.5. High EC (\>2 dS/m) indicates salinity stress. High ESP (\>15%) causes soil structure degradation. Low C:N ratio (\<15) favors rapid nitrogen release.

```{r soil_plots}
soil <- read_csv("soil_data.csv", show_col_types = FALSE)

vars_soil <- c("pH", "organic_matter", "total_nitrogen",
               "cec", "clay", "sand", "silt")

p_soil <- soil |>
  pivot_longer(cols = all_of(vars_soil),
               names_to  = "property",
               values_to = "value") |>
  mutate(property = factor(property, levels = vars_soil)) |>
  ggplot(aes(x = value,
             y = reorder(oblast, value),
             fill = property)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  facet_wrap(~property, scales = "free_x", nrow = 2) +
  scale_fill_manual(
    values = c("#2c7bb6", "#d7191c", "#1a9641",
               "#fdae61", "#abd9e9", "#756bb1", "#a6bddb")
  ) +
  labs(
    title   = "Soil properties by oblast",
    caption = "Source: soil_data.csv"
  ) +
  theme_small_multiple(base_size = 8) +
  theme(
    axis.text.y    = element_text(size = 6),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3)
  )


print(p_soil)
```

------------------------------------------------------------------------

### 2.4 Correlation structure of all variables

In exploratory research it is important to check correlations not only among variables we expect to be related, but across all pairs — including crop variables, weather metrics, and soil moisture together. Unexpected correlations can reveal data artefacts, suggest omitted mechanisms, or warn about multicollinearity before model estimation.

```{r build-full-panel}
library(tidyverse)
library(corrplot)
library(GGally)
library(car)

# ── Region name lookup (Latin weather ↔ Cyrillic crop data) ──
region_lookup <- tribble(
  ~region_lat,        ~region_cyr,
  "Vinnytsia",        "Вінницька",
  "Volyn",            "Волинська",
  "Dnipropetrovsk",   "Дніпропетровська",
  "Donetsk",          "Донецька",
  "Zhytomyr",         "Житомирська",
  "Zakarpattia",      "Закарпатська",
  "Zaporizhzhia",     "Запорізька",
  "Ivano-Frankivsk",  "Івано-Франківська",
  "Kyiv",             "Київська",
  "Kirovohrad",       "Кіровоградська",
  "Luhansk",          "Луганська",
  "Lviv",             "Львівська",
  "Mykolaiv",         "Миколаївська",
  "Odesa",            "Одеська",
  "Poltava",          "Полтавська",
  "Rivne",            "Рівненська",
  "Sumy",             "Сумська",
  "Ternopil",         "Тернопільська",
  "Kharkiv",          "Харківська",
  "Kherson",          "Херсонська",
  "Khmelnytskyi",     "Хмельницька",
  "Cherkasy",         "Черкаська",
  "Chernivtsi",       "Чернівецька",
  "Chernihiv",        "Чернігівська"
)

# ── Weather annual aggregation including soil moisture ────────
weather_annual_full <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - 5,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20, t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0, t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    # Thermal stress metrics
    days_hot_crit    = sum(days_hot_crit,       na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit,      na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,        na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,        na.rm = TRUE),
    # Raw temperature levels
    mean_tmax        = mean(temperature_2m_max, na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min, na.rm = TRUE),
    mean_temp        = mean(t_mean,             na.rm = TRUE),
    # Water and energy
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    total_radiation  = sum(shortwave_radiation_sum,      na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    mean_dewpoint    = mean(dewpoint_2m_mean,            na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(year >= 2015)

# ── Soil moisture aggregation (same logic as temperature) ─────
soil_moisture_agg <- soil_moisture |>
  mutate(
    drought_day = soil_moisture_0_7cm_pct < 20,
    waterlog_day = soil_moisture_0_7cm_pct > 85
  ) |>
  group_by(region, year) |>
  summarise(
    mean_soil_moisture_pct = mean(soil_moisture_0_7cm_pct, na.rm = TRUE),
    days_drought = sum(drought_day, na.rm = TRUE),
    days_waterlog = sum(waterlog_day, na.rm = TRUE),
    .groups = "drop"
  )

# ── Harvest ratio ─────────────────────────────────────────────
harv_ratio_df <- harv |>
  left_join(sown, by = c("region", "year")) |>
  mutate(harv_ratio = pmin(harv_ha / sown_ha, 1.05)) |>
  select(region, year, harv_ratio)

# ── Join everything on Cyrillic region names ──────────────────
panel_full <- yield |>
  left_join(harv_ratio_df, by = c("region", "year")) |>
  left_join(region_lookup, by = c("region" = "region_cyr")) |>
  left_join(weather_annual_full,
            by = c("region_lat" = "region", "year")) |>
  left_join(soil_moisture_agg,
            by = c("region_lat" = "region", "year")) |>
  left_join(soil |> rename(region_cyr = oblast),
            by = c("region" = "region_cyr")) |>
  filter(!is.na(yield_cpha))

cat("Panel rows:", nrow(panel_full), "\n")
cat("Regions matched:", n_distinct(panel_full$region[!is.na(panel_full$mean_temp)]), "\n")
```

```{r collinearity, fig.width = 10, fig.height = 9}
# ── Regressor set: includes all variables to prove redundancy ──
regressors <- c(
  "days_hot_crit",
  "days_cold_crit", 
  "sum_heat_noncrit",
  "sum_cold_noncrit",
  "total_precip",
  "total_et0",
  "total_radiation",
  "mean_wind",
  "mean_dewpoint",
  "mean_soil_moisture_pct",
  "days_drought",
  "harv_ratio"
)

# Keep only columns that exist
regressors <- intersect(regressors, names(panel_full))

reg_df <- panel_full |>
  select(all_of(regressors)) |>
  drop_na()

# Also create a version with mean_temp for correlation checks
reg_df_with_temp <- panel_full |>
  select(all_of(c(regressors, "mean_temp"))) |>
  drop_na()

# ── Display labels ────────────────────────────────────────────
reg_labels <- c(
  days_hot_crit    = "Hot critical days",
  days_cold_crit   = "Cold critical days",
  sum_heat_noncrit = "Accum. heat",
  sum_cold_noncrit = "Accum. cold",
  total_precip     = "Precipitation",
  total_et0        = "ET0",
  total_radiation  = "Solar radiation",
  mean_wind        = "Wind gusts",
  mean_dewpoint    = "Dewpoint",
  mean_soil_moisture_pct = "Mean soil moisture",
  days_drought     = "Drought days",
  harv_ratio       = "Harvest ratio"
)

reg_labels <- reg_labels[regressors]

# ── Correlation matrix among regressors ────────────────────
cor_reg <- cor(reg_df, use = "pairwise.complete.obs")
rownames(cor_reg) <- colnames(cor_reg) <- reg_labels

par(bg = "white")
corrplot(
  cor_reg,
  method      = "color",
  type        = "upper",
  order       = "hclust",
  tl.col      = "grey15",
  tl.srt      = 45,
  tl.cex      = 0.85,
  cl.cex      = 0.75,
  addCoef.col = "grey10",
  number.cex  = 0.7,
  col         = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(200),
  title       = "Correlation matrix among regressors",
  mar         = c(0, 0, 2, 0)
)
```

```{r vif}
# ── VIF for the full regressor set ──
vif_formula <- as.formula(
  paste("yield_cpha ~", paste(regressors, collapse = " + "))
)

vif_data <- panel_full |>
  select(yield_cpha, all_of(regressors)) |>
  drop_na()

vif_model  <- lm(vif_formula, data = vif_data)
vif_values <- vif(vif_model)

vif_df <- tibble(
  variable = names(vif_values),
  VIF      = vif_values
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High — must address",
      VIF > 5  ~ "Moderate — monitor",
      TRUE     ~ "Acceptable"
    )
  ) |>
  arrange(desc(VIF))

# Print VIF table
knitr::kable(
  vif_df |> select(label, VIF, concern),
  digits  = 2,
  col.names = c("Variable", "VIF", "Concern"),
  caption = "VIF — full set (all weather variables included)"
)

# Print key correlations (using reg_df_with_temp which already has all aligned data)
cat("\n=== Evidence for Removing Variables ===\n")
cat("Correlation ET0 vs Solar radiation:", 
    round(cor(reg_df$total_et0, reg_df$total_radiation, use = "complete.obs"), 2), "\n")
cat("Correlation Dewpoint vs Mean temperature:", 
    round(cor(reg_df_with_temp$mean_dewpoint, reg_df_with_temp$mean_temp, use = "complete.obs"), 2), "\n")
cat("Correlation ET0 vs Hot days:", 
    round(cor(reg_df$total_et0, reg_df$days_hot_crit, use = "complete.obs"), 2), "\n")
cat("Correlation Solar radiation vs Hot days:", 
    round(cor(reg_df$total_radiation, reg_df$days_hot_crit, use = "complete.obs"), 2), "\n")

cat("\n=== Soil Moisture Correlations ===\n")
cat("Mean soil moisture vs Precipitation:", 
    round(cor(reg_df$mean_soil_moisture_pct, reg_df$total_precip, use = "complete.obs"), 2), "\n")
cat("Mean soil moisture vs ET0:", 
    round(cor(reg_df$mean_soil_moisture_pct, reg_df$total_et0, use = "complete.obs"), 2), "\n")
cat("Drought days vs Hot days:", 
    round(cor(reg_df$days_drought, reg_df$days_hot_crit, use = "complete.obs"), 2), "\n")
cat("Drought days vs Precipitation:", 
    round(cor(reg_df$days_drought, reg_df$total_precip, use = "complete.obs"), 2), "\n")
```

```{r vif-reduced, fig.width = 8, fig.height = 5}
# ── VIF after removing redundant variables: radiation, dewpoint, and ET0 ──
# Keeping only precipitation as the water/energy representative
regressors_reduced <- regressors[!regressors %in% c("total_radiation", "mean_dewpoint", "total_et0")]

vif_formula2 <- as.formula(
  paste("yield_cpha ~", paste(regressors_reduced, collapse = " + "))
)

vif_model2  <- lm(vif_formula2,
                  data = panel_full |>
                    select(yield_cpha, all_of(regressors_reduced)) |>
                    drop_na())
vif_values2 <- vif(vif_model2)

vif_df2 <- tibble(
  variable = names(vif_values2),
  VIF      = vif_values2
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High",
      VIF > 5  ~ "Moderate",
      TRUE     ~ "Acceptable"
    )
  ) |>
  arrange(desc(VIF))

knitr::kable(
  vif_df2 |> select(label, VIF, concern),
  digits  = 2,
  col.names = c("Variable", "VIF", "Concern"),
  caption = "VIF after removing solar radiation, dewpoint, and ET0"
)

cat("\n=== Model Improvement Summary ===\n")
cat("Maximum VIF before removal:", round(max(vif_df$VIF), 2), "\n")
cat("Maximum VIF after removal:", round(max(vif_df2$VIF), 2), "\n")
cat("Variables removed:\n")
cat("  - total_radiation (correlated with ET0 at r > 0.85)\n")
cat("  - mean_dewpoint (correlated with temperature at r > 0.70)\n")
cat("  - total_et0 (correlated with hot days and radiation)\n")
cat("Remaining water-related variable: total_precip (precipitation)\n")
```

```{r vif-plot, fig.width = 8, fig.height = 5}
# VIF bar chart for reduced model
ggplot(vif_df2,
       aes(x = VIF,
           y = fct_reorder(label, VIF),
           fill = concern)) +
  geom_col(width = 0.68, alpha = 0.88) +
  geom_text(aes(label = round(VIF, 1)),
            hjust = -0.15, size = 3, colour = "grey25") +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#d7191c", linewidth = 0.5) +
  scale_fill_manual(values = c(
    "Acceptable" = "#1a9641",
    "Moderate"   = "#fdae61",
    "High"       = "#d7191c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title   = "VIF after removing redundant variables",
    subtitle = "Solar radiation, dewpoint, and ET0 removed (kept precipitation only)",
    x       = "VIF", y = NULL, fill = NULL,
    caption = "Dashed line = VIF 5, Red line = VIF 10"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    legend.position = "top"
  )
```

### Conclusions from Correlation and VIF Analysis

The correlation and VIF analysis reveals several important findings for model specification:

**1. Severe Multicollinearity in Initial Variable Set**

The full model with all weather variables exhibited extreme multicollinearity, with a maximum VIF of **70.25** — far exceeding the conventional threshold of 5-10. This confirms our expectation that variables such as solar radiation, ET0, and dewpoint are largely redundant given the information already captured by temperature metrics and precipitation.

**2. Successful Reduction of Redundant Variables**

After removing solar radiation, dewpoint, and ET0, the maximum VIF dropped dramatically from 70.25 to **6.91** — a tenfold reduction. All remaining variables now fall within acceptable or moderate VIF ranges, indicating that multicollinearity is no longer a concern for model estimation.

**3. Soil Moisture Variables Require Caution**

Both `Drought days` (VIF = 6.91) and `Mean soil moisture` (VIF = 5.95) show moderate multicollinearity. This is expected, as drought conditions are naturally correlated with high temperatures and low precipitation. However, their VIF values remain below 10, meaning they can be retained in the model, but their standard errors will be somewhat inflated. The fact that drought days shows higher VIF than mean soil moisture suggests that the extreme-event count captures information more closely related to other weather variables than the annual average does.

**4. Acceptable Variables for Final Model**

The following variables show good VIF values (all below 2) and can be confidently included: - Temperature stress metrics (hot days, cold days, accumulated heat, accumulated cold) - Precipitation - Wind gusts - Harvest ratio

**5. Implications for Model Specification**

Based on this analysis, our final modelling strategy will: - **Exclude** solar radiation, dewpoint, and ET0 (redundant with temperature metrics) - **Retain** precipitation as the sole water-related variable from standard weather data - **Include** soil moisture metrics (mean and drought days) despite moderate multicollinearity, as they capture unique information about plant-available water that precipitation alone cannot provide - **Monitor** standard errors for soil moisture coefficients, recognising they may be slightly inflated

This parsimonious specification balances theoretical completeness with statistical efficiency, avoiding the inefficiencies of multicollinearity while preserving the core mechanisms hypothesised to affect wheat yield.

------------------------------------------------------------------------

# 3. Modelling

This section builds a sequence of OLS regression models with explicit region dummy variables (Київська oblast as the reference category) to analyse the impact of weather on winter wheat yield in Ukraine. Based on the correlation analysis from Section 2.4, we exclude solar radiation, dewpoint, and ET0 due to high multicollinearity with temperature metrics. Precipitation is retained as the sole water-related variable from standard weather data, alongside soil moisture metrics which capture plant-available water.

We proceed in three parts: stationarity and trend (3.1), weather window and aggregation (3.2), and soil variables (3.3).

**Note on standard errors:** All models use plain OLS standard errors. This assumes homoskedasticity and no serial correlation in the error term — an assumption that may not hold perfectly in panel data. Results should be interpreted with this limitation in mind.

------------------------------------------------------------------------

## 3.0 Setup and panel construction

```{r part3-libs}
library(tidyverse)
library(lubridate)
library(lmtest)
library(sandwich)
library(lme4)
library(modelsummary)
library(patchwork)
library(tseries)
library(car)
library(corrplot)
library(fastDummies)
library(multcomp)
```

```{r part3-setup}
# ── Region lookup ─────────────────────────────────────────────
region_lookup <- tribble(
  ~region_lat,        ~region_cyr,          ~region_soil,
  "Vinnytsia",        "Вінницька",          "Vinnytska",
  "Volyn",            "Волинська",          "Volynska",
  "Dnipropetrovsk",   "Дніпропетровська",   "Dnipropetrovska",
  "Donetsk",          "Донецька",           "Donetska",
  "Zhytomyr",         "Житомирська",        "Zhytomyrska",
  "Zakarpattia",      "Закарпатська",       "Zakarpatska",
  "Zaporizhzhia",     "Запорізька",         "Zaporizka",
  "Ivano-Frankivsk",  "Івано-Франківська",  "Ivano-Frankivska",
  "Kyiv",             "Київська",           "Kyivska",
  "Kirovohrad",       "Кіровоградська",     "Kirovohradska",
  "Luhansk",          "Луганська",          "Luhanska",
  "Lviv",             "Львівська",          "Lvivska",
  "Mykolaiv",         "Миколаївська",       "Mykolaivska",
  "Odesa",            "Одеська",            "Odeska",
  "Poltava",          "Полтавська",         "Poltavska",
  "Rivne",            "Рівненська",         "Rivnenska",
  "Sumy",             "Сумська",            "Sumska",
  "Ternopil",         "Тернопільська",      "Ternopilska",
  "Kharkiv",          "Харківська",         "Kharkivska",
  "Kherson",          "Херсонська",         "Khersonska",
  "Khmelnytskyi",     "Хмельницька",        "Khmelnytska",
  "Cherkasy",         "Черкаська",          "Cherkaska",
  "Chernivtsi",       "Чернівецька",        "Chernivetska",
  "Chernihiv",        "Чернігівська",       "Chernihivska"
)

# ── Soil ──────────────────────────────────────────────────────
soil <- read_csv("soil_data.csv", show_col_types = FALSE)
names(soil)[names(soil) == "oblast"] <- "region"

# ── Weather aggregation helper (without ET0, radiation, dewpoint) ──
# Based on correlation analysis, we keep only temperature metrics, precipitation, and wind
BASE_TEMP <- 5

agg_weather_slice <- function(df, suffix = "") {
  result <- df %>%
    mutate(
      t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
      t_dev          = t_mean - BASE_TEMP,
      days_hot_crit  = as.integer(t_dev > 20),
      days_cold_crit = as.integer(t_dev < -15),
      heat_pos       = if_else(t_dev > 0, t_dev, 0),
      cold_neg       = if_else(t_dev < 0, t_dev, 0)
    ) %>%
    group_by(region, year) %>%
    summarise(
      days_hot_crit  = sum(days_hot_crit,         na.rm = TRUE),
      days_cold_crit = sum(days_cold_crit,        na.rm = TRUE),
      sum_heat_pos   = sum(heat_pos,              na.rm = TRUE),
      sum_cold_neg   = sum(cold_neg,              na.rm = TRUE),
      total_precip   = sum(precipitation_sum,     na.rm = TRUE),
      mean_wind      = mean(wind_gusts_10m_max,   na.rm = TRUE),
      .groups = "drop"
    )
  
  if (suffix != "") {
    result <- result %>%
      rename_with(~ paste0(.x, suffix), -c(region, year))
  }
  
  return(result)
}

# ── Load weather ──────────────────────────────────────────────
weather_raw <- read_csv("daily_weather.csv", show_col_types = FALSE) %>%
  mutate(date = as_date(date), year = year(date)) %>%
  filter(year >= 2015, year <= 2024)

weather_tagged <- weather_raw %>%
  mutate(
    cal_month  = month(date),
    cal_day    = day(date),
    harvest_yr = if_else(cal_month >= 9, year + 1L, year)
  )

# ── Seasonal slices ──
weather_full <- weather_raw %>%
  agg_weather_slice(suffix = "")

weather_aug <- weather_raw %>%
  filter(month(date) == 8) %>%
  agg_weather_slice(suffix = "_aug")

weather_gs <- weather_tagged %>%
  filter((cal_month == 9 & cal_day >= 10) |
         cal_month %in% c(10, 11, 12, 1, 2, 3, 4, 5, 6, 7)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_gs")

weather_early <- weather_tagged %>%
  filter(cal_month %in% c(10, 11, 12, 1, 2, 3, 4) |
         (cal_month == 5 & cal_day <= 15)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_early")

weather_late <- weather_tagged %>%
  filter((cal_month == 5 & cal_day >= 16) | cal_month %in% c(6, 7)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_late")

weather_w1 <- weather_tagged %>%
  filter(cal_month %in% c(10, 11, 12, 1)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_w1")

weather_w2 <- weather_tagged %>%
  filter(cal_month %in% c(2, 3, 4)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_w2")

weather_w3 <- weather_tagged %>%
  filter((cal_month == 5 & cal_day >= 16) | cal_month %in% c(6, 7)) %>%
  dplyr::select(-c(year, cal_month, cal_day)) %>%
  rename(year = harvest_yr) %>%
  agg_weather_slice(suffix = "_w3")

# ── Load soil moisture ─────────────────────────────────────────
soil_moisture <- read_csv("soil_moisture.csv", show_col_types = FALSE) %>%
  mutate(date = as_date(date), year = year(date)) %>%
  filter(year >= 2015, year <= 2024)

soil_moisture_agg <- soil_moisture %>%
  mutate(
    drought_day = soil_moisture_0_7cm_pct < 20,
    waterlog_day = soil_moisture_0_7cm_pct > 85
  ) %>%
  group_by(region, year) %>%
  summarise(
    mean_soil_moisture_pct = mean(soil_moisture_0_7cm_pct, na.rm = TRUE),
    days_drought = sum(drought_day, na.rm = TRUE),
    .groups = "drop"
  )

# ── Load crop data ────────────────────────────────────────────
META_COLS <- c("indicator", "region", "unit", "scale", "decimals", "special", "note_ts", "note_obs")
years <- as.character(2015:2025)

read_panel <- function(file, value_name) {
  read_csv(file, col_names = c(META_COLS, years), skip = 1, show_col_types = FALSE) %>%
    dplyr::select(region, all_of(years)) %>%
    pivot_longer(cols = all_of(years), names_to = "year", values_to = value_name) %>%
    mutate(year = as.integer(year), !!value_name := suppressWarnings(as.numeric(.data[[value_name]]))) %>%
    filter(region != "Україна")
}

yield <- read_panel("crop_capacity.csv", "yield_cpha")
sown <- read_panel("sowing_area.csv", "sown_kha") %>% mutate(sown_ha = sown_kha * 1000) %>% dplyr::select(-sown_kha)
harv <- read_panel("harvested_area.csv", "harv_kha") %>% mutate(harv_ha = harv_kha * 1000) %>% dplyr::select(-harv_kha)

occupied_oblasts <- c("Донецька", "Луганська", "Запорізька", "Херсонська")

# ── Build panel_base ──────────────────────────────────────────
panel_base <- yield %>%
  left_join(sown %>% dplyr::select(region, year, sown_ha), by = c("region", "year")) %>%
  left_join(harv %>% dplyr::select(region, year, harv_ha), by = c("region", "year")) %>%
  left_join(region_lookup, by = c("region" = "region_cyr")) %>%
  left_join(weather_full,  by = c("region_lat" = "region", "year")) %>%
  left_join(weather_aug,   by = c("region_lat" = "region", "year")) %>%
  left_join(weather_gs,    by = c("region_lat" = "region", "year")) %>%
  left_join(weather_early, by = c("region_lat" = "region", "year")) %>%
  left_join(weather_late,  by = c("region_lat" = "region", "year")) %>%
  left_join(weather_w1,    by = c("region_lat" = "region", "year")) %>%
  left_join(weather_w2,    by = c("region_lat" = "region", "year")) %>%
  left_join(weather_w3,    by = c("region_lat" = "region", "year")) %>%
  left_join(soil_moisture_agg, by = c("region_lat" = "region", "year")) %>%
  left_join(soil,          by = "region") %>%
  mutate(
    war_east  = as.integer(region %in% occupied_oblasts & year == 2022),
    log_yield = log(yield_cpha),
    log_sown  = log(sown_ha + 1),
    t         = year - min(year),
    region_id = relevel(as.factor(region), ref = "Київська")
  ) %>%
  filter(!is.na(yield_cpha), !is.na(days_hot_crit))

# ── Calculate loss_lag ────────────────────────────────────────
loss_rate_raw <- panel_base %>%
  mutate(
    loss_rate = pmax((sown_ha - harv_ha) / sown_ha, 0),
    loss_rate = pmin(loss_rate, 0.5)
  )

loss_lag <- loss_rate_raw %>%
  dplyr::select(region, year, loss_rate) %>%
  mutate(year = year + 1, loss_lag = loss_rate) %>%
  dplyr::select(-loss_rate)

panel_base <- panel_base %>%
  left_join(loss_lag, by = c("region", "year"))

# ── Fix region names and create dummies ───────────────────────
panel_base$region <- gsub("-", "_", panel_base$region)
panel_base$region_id <- relevel(as.factor(panel_base$region), ref = "Київська")

region_dummies_raw <- model.matrix(~ region_id, data = panel_base) %>%
  as.data.frame() %>%
  dplyr::select(-`(Intercept)`)

names(region_dummies_raw) <- make.names(names(region_dummies_raw))
panel_base <- bind_cols(panel_base, region_dummies_raw)
dummy_cols <- names(region_dummies_raw)

# ── Verification ──────────────────────────────────────────────
cat("Panel rows:", nrow(panel_base), "\n")
cat("Regions:  ", n_distinct(panel_base$region), "\n")
cat("Years:    ", n_distinct(panel_base$year), "\n")
cat("War obs:  ", sum(panel_base$war_east, na.rm = TRUE), "\n")
cat("Dummy columns created:", length(dummy_cols), "\n")
```

------------------------------------------------------------------------

# Part 1 — Stationarity and trend

## 3.1.1 ADF tests on yield per oblast

```{r adf-yield}
adf_results <- panel_base |>
  group_by(region) |>
  arrange(year) |>
  summarise(
    adf_stat  = tryCatch(adf.test(yield_cpha, k = 1)$statistic, error = function(e) NA_real_),
    adf_pval  = tryCatch(adf.test(yield_cpha, k = 1)$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(
    stationary = case_when(
      adf_pval < 0.01 ~ "Yes ***",
      adf_pval < 0.05 ~ "Yes **",
      adf_pval < 0.10 ~ "Yes *",
      TRUE            ~ "No"
    )
  ) |>
  arrange(adf_pval)

knitr::kable(adf_results, digits = 3, caption = "ADF test results for yield per oblast (k=1 lag)")
```

## 3.1.2 Baseline model (no trend)

```{r baseline-model}
weather_vars_full <- c("days_hot_crit", "days_cold_crit", "sum_heat_pos", "sum_cold_neg", 
                       "total_precip", "mean_wind", "mean_soil_moisture_pct", "days_drought")

fml_base <- as.formula(paste(
  "yield_cpha ~",
  paste(c(weather_vars_full, "log_sown", "war_east", dummy_cols), collapse = " + ")
))

m_base <- lm(fml_base, data = panel_base)
summary(m_base)
```

## 3.1.3 Common time trend

```{r common-trend}
fml_trend <- as.formula(paste(
  "yield_cpha ~",
  paste(c(weather_vars_full, "log_sown", "war_east", "t", dummy_cols), collapse = " + ")
))

m_trend <- lm(fml_trend, data = panel_base)
summary(m_trend)
```

## 3.1.4 Region-specific trends

```{r region-trends}
region_t_cols <- paste0(dummy_cols, "_t")
for (col in dummy_cols) {
  panel_base[[paste0(col, "_t")]] <- panel_base[[col]] * panel_base$t
}

fml_region_trend <- as.formula(paste(
  "yield_cpha ~",
  paste(c(weather_vars_full, "log_sown", "war_east", "t", dummy_cols, region_t_cols), collapse = " + ")
))

m_region_trend <- lm(fml_region_trend, data = panel_base)
summary(m_region_trend)
```

## 3.1.5 Adding lagged loss rate

```{r lagged-loss}
panel_lag <- panel_base |> filter(!is.na(loss_lag))

fml_lag <- as.formula(paste(
  "yield_cpha ~",
  paste(c(weather_vars_full, "log_sown", "war_east", "loss_lag", dummy_cols), collapse = " + ")
))

m_lag <- lm(fml_lag, data = panel_lag)
summary(m_lag)
```

## 3.1.6 ADF on residuals

```{r adf-residuals}
panel_base$resid_base <- residuals(m_base)
panel_base$resid_trend <- residuals(m_trend)

adf_resid <- panel_base |>
  group_by(region) |>
  arrange(year) |>
  summarise(
    adf_base_p  = tryCatch(adf.test(resid_base, k = 1)$p.value, error = function(e) NA_real_),
    adf_trend_p = tryCatch(adf.test(resid_trend, k = 1)$p.value, error = function(e) NA_real_),
    .groups = "drop"
  )

knitr::kable(adf_resid, digits = 3, caption = "ADF on model residuals per oblast")
```

## 3.1.7 Part 1 comparison table

```{r part1-comparison}
part1_table <- tibble(
  Model = c("Baseline (no trend)", "+ Common trend (t)", "+ Region trends (t × region)", "+ Lagged loss rate"),
  N = c(nobs(m_base), nobs(m_trend), nobs(m_region_trend), nobs(m_lag)),
  Adj_R2 = c(summary(m_base)$adj.r.squared, summary(m_trend)$adj.r.squared,
             summary(m_region_trend)$adj.r.squared, summary(m_lag)$adj.r.squared),
  AIC = c(AIC(m_base), AIC(m_trend), AIC(m_region_trend), AIC(m_lag)),
  BIC = c(BIC(m_base), BIC(m_trend), BIC(m_region_trend), BIC(m_lag))
)

knitr::kable(part1_table, digits = 3, caption = "Part 1 model comparison")
```

------------------------------------------------------------------------

# Part 2 — Weather window and aggregation

## 3.2.0 Choosing best Part 1 specification

```{r best-part1}
adj_r2_part1 <- c(summary(m_base)$adj.r.squared, summary(m_trend)$adj.r.squared, summary(m_region_trend)$adj.r.squared)
best_part1 <- c("none", "common", "region")[which.max(adj_r2_part1)]
cat("Best trend specification:", best_part1, "\n")

trend_controls <- switch(best_part1,
  none   = character(0),
  common = "t",
  region = c("t", region_t_cols)
)

base_controls <- c("log_sown", "war_east", dummy_cols)
all_controls  <- c(base_controls, trend_controls)
```

## 3.2.1 Full-year model (M_full)

```{r m-full}
fml_full <- as.formula(paste("yield_cpha ~", paste(c(weather_vars_full, all_controls), collapse = " + ")))
m_full <- lm(fml_full, data = panel_base)
summary(m_full)
```

## 3.2.2 Growing season model (M_gs)

```{r m-gs}
weather_vars_gs <- c("days_hot_crit_gs", "days_cold_crit_gs", "sum_heat_pos_gs", "sum_cold_neg_gs",
                     "total_precip_gs", "mean_wind_gs", "mean_soil_moisture_pct", "days_drought")

fml_gs <- as.formula(paste("yield_cpha ~", paste(c(weather_vars_gs, all_controls), collapse = " + ")))
m_gs <- lm(fml_gs, data = panel_base)
summary(m_gs)
```

## 3.2.3 Two-split model (M_split2)

```{r m-split2}
weather_vars_split2 <- c(
  "days_hot_crit_early", "days_cold_crit_early", "sum_heat_pos_early", "sum_cold_neg_early",
  "total_precip_early", "mean_wind_early",
  "days_hot_crit_late", "days_cold_crit_late", "sum_heat_pos_late", "sum_cold_neg_late",
  "total_precip_late", "mean_wind_late"
)

fml_split2 <- as.formula(paste("yield_cpha ~", paste(c(weather_vars_split2, all_controls), collapse = " + ")))
m_split2 <- lm(fml_split2, data = panel_base)
summary(m_split2)
```

## 3.2.4 Three-split model (M_split3)

```{r m-split3}
weather_vars_split3 <- c(
  "days_hot_crit_w1", "days_cold_crit_w1", "sum_heat_pos_w1", "sum_cold_neg_w1", "total_precip_w1", "mean_wind_w1",
  "days_hot_crit_w2", "days_cold_crit_w2", "sum_heat_pos_w2", "sum_cold_neg_w2", "total_precip_w2", "mean_wind_w2",
  "days_hot_crit_w3", "days_cold_crit_w3", "sum_heat_pos_w3", "sum_cold_neg_w3", "total_precip_w3", "mean_wind_w3"
)

fml_split3 <- as.formula(paste("yield_cpha ~", paste(c(weather_vars_split3, all_controls), collapse = " + ")))
m_split3 <- lm(fml_split3, data = panel_base)
summary(m_split3)
```

## 3.2.5 Part 2 comparison table

```{r part2-comparison}
part2_table <- tibble(
  Model = c("M_full (full year)", "M_gs (growing season)", "M_split2 (early + late)", "M_split3 (3 windows)"),
  N = c(nobs(m_full), nobs(m_gs), nobs(m_split2), nobs(m_split3)),
  Adj_R2 = c(summary(m_full)$adj.r.squared, summary(m_gs)$adj.r.squared,
             summary(m_split2)$adj.r.squared, summary(m_split3)$adj.r.squared),
  AIC = c(AIC(m_full), AIC(m_gs), AIC(m_split2), AIC(m_split3)),
  BIC = c(BIC(m_full), BIC(m_gs), BIC(m_split2), BIC(m_split3))
)

knitr::kable(part2_table, digits = 3, caption = "Part 2 model comparison")
```

------------------------------------------------------------------------

# Part 3 — Soil variables

## 3.3.0 Theoretical background

### Why region dummies eliminate time-invariant variables

In a model with explicit region dummy variables, each dummy captures the full average level of yield for that oblast, absorbing all time-invariant characteristics — including soil properties. Formally, for any time-invariant variable $z_i$ (such as soil organic matter), its value is identical across all $T$ observations belonging to region $i$. This means $z_i$ is a linear combination of the region dummies:

$$z_i = \sum_{j} z_j \cdot \mathbf{1}[\text{region} = j]$$

Including $z_i$ alongside the region dummies therefore creates **perfect multicollinearity** — the design matrix is rank-deficient and the soil coefficient $\gamma$ cannot be separately identified from the region intercepts. R will silently drop the collinear columns.

This is not a failure of the model — it is a feature. The region dummies are doing exactly what we want: controlling for all stable regional differences including soil, climate, infrastructure, and management culture. The cost is that we cannot estimate the direct effect of soil on yield.

### Random effects model

The random effects (RE) model offers a way to recover soil coefficients by treating region-level unobservables as random draws from a distribution rather than fixed parameters to be estimated:

$$y_{it} = \mathbf{x}_{it}'\boldsymbol{\beta} + \mathbf{z}_i'\boldsymbol{\gamma} + u_i + \varepsilon_{it}$$

where $u_i \sim \mathcal{N}(0, \sigma_u^2)$ is the region-level random effect and $\varepsilon_{it} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$ is the idiosyncratic error. The key assumption is **orthogonality**: $\text{Cov}(u_i, \mathbf{x}_{it}) = 0$ and $\text{Cov}(u_i, \mathbf{z}_i) = 0$. Under this assumption, $\boldsymbol{\gamma}$ — the soil coefficients — is consistently estimated.

However, this assumption is likely violated in our setting. Soil quality is geographically correlated with climate: southern oblasts tend to have both chernozem soils (high organic matter) and hotter, drier weather. This means $\text{Cov}(u_i, \mathbf{x}_{it}) \neq 0$, and RE estimates of soil effects will absorb some of the omitted weather-climate variation, biasing $\boldsymbol{\gamma}$.

### The Mundlak-Chamberlain device

The Mundlak (1978) and Chamberlain (1982) approach offers a middle ground between region dummies and pure RE. It augments the RE model by including the within-region means $\bar{\mathbf{x}}_i$ of all time-varying regressors as additional controls:

$$y_{it} = \mathbf{x}_{it}'\boldsymbol{\beta} + \mathbf{z}_i'\boldsymbol{\gamma} + \bar{\mathbf{x}}_i'\boldsymbol{\lambda} + a_i + \varepsilon_{it}$$

where $a_i$ is now assumed orthogonal to both $\mathbf{x}_{it} - \bar{\mathbf{x}}_i$ and $\mathbf{z}_i$. Three key results follow:

1.  The $\boldsymbol{\beta}$ estimates are numerically identical to the within (FE/dummy) estimator — time-varying weather effects are consistently estimated regardless of whether RE assumptions hold
2.  The $\boldsymbol{\gamma}$ estimates recover soil coefficients **conditional on the long-run average weather environment** of each oblast, captured by $\bar{\mathbf{x}}_i$. This controls for the climate-soil correlation that would otherwise bias RE
3.  A joint Wald test on $\boldsymbol{\lambda} = \mathbf{0}$ is a Hausman-type specification test: if the Mundlak terms are jointly insignificant, RE orthogonality holds and plain RE is valid; if rejected, Mundlak is preferred

In our context, $\bar{\mathbf{x}}_i$ captures the long-run average weather of each oblast — essentially its climate. Including these means allows soil coefficients to be interpreted as the effect of soil quality **conditional on local climate**, which is the variation we want to identify.

### Soil $\times$ weather interactions under region dummies

A third approach identifies soil effects through their **interaction** with time-varying weather variables. For example:

$$y_{it} = \dots + \beta_1 \cdot \text{hot\_days}_{it} + \beta_2 \cdot \text{OM}_i + \beta_3 \cdot (\text{hot\_days}_{it} \times \text{OM}_i) + \text{region dummies} + \varepsilon_{it}$$

The main effect $\beta_2$ of $\text{OM}_i$ is collinear with the region dummies and is dropped. However, the interaction term $\beta_3$ **is identified** because $\text{hot\_days}_{it} \times \text{OM}_i$ varies over time (hot days change each year) even though $\text{OM}_i$ is fixed. This coefficient answers a more policy-relevant question: **does having higher organic matter change how sensitive yield is to heat or drought stress?** If $\beta_3 > 0$ for `organic_matter × days_hot_crit`, then oblasts with richer soils suffer less yield loss per additional hot day — a direct argument for soil investment as climate adaptation.

------------------------------------------------------------------------

## 3.3.1 Random effects model

We estimate a random effects model using `lme4::lmer` with soil variables as direct regressors alongside the best weather specification from Part 2. We use the growing season weather variables (`_gs`) as the weather block, since the Part 2 comparison will determine the best window — the growing season model is a reasonable default and avoids overfitting from the split models.

Note that soil moisture variables (`mean_soil_moisture_pct`, `days_drought`) are time-varying and are included as normal weather regressors. Only the static soil properties from `soil_data.csv` (pH, organic matter, etc.) face the collinearity problem and require the RE/Mundlak approach.

```{r}
oblast_map <- c(
  "Vinnytska"        = "Вінницька",
  "Volynska"         = "Волинська",
  "Dnipropetrovska"  = "Дніпропетровська",
  "Donetska"         = "Донецька",
  "Zhytomyrska"      = "Житомирська",
  "Zakarpatska"      = "Закарпатська",
  "Zaporizka"        = "Запорізька",
  "Ivano_Frankivska" = "Івано_Франківська",
  "Kyivska"          = "Київська",
  "Kirovohradska"    = "Кіровоградська",
  "Luhanska"         = "Луганська",
  "Lvivska"          = "Львівська",
  "Mykolaivska"      = "Миколаївська",
  "Odeska"           = "Одеська",
  "Poltavska"        = "Полтавська",
  "Rivnenska"        = "Рівненська",
  "Sumska"           = "Сумська",
  "Ternopilska"      = "Тернопільська",
  "Kharkivska"       = "Харківська",
  "Khersonska"       = "Херсонська",
  "Khmelnytska"      = "Хмельницька",
  "Cherkaska"        = "Черкаська",
  "Chernivetska"     = "Чернівецька",
  "Chernihivska"     = "Чернігівська"
)

soil_raw <- read_csv("soil_data.csv", show_col_types = FALSE) |>
  mutate(region = oblast_map[oblast]) |>
  dplyr::select(-oblast)

panel_base <- panel_base |>
  left_join(soil_raw, by = "region")

panel_base$region <- trimws(panel_base$region)
soil_raw$region   <- trimws(soil_raw$region)

panel_base$region <- enc2utf8(panel_base$region)
soil_raw$region   <- enc2utf8(soil_raw$region)

panel_base <- panel_base |>
  left_join(soil_raw, by = "region")

names(panel_base)
```


```{r re-model}
soil_vars <- c("pH", "organic_matter", "total_nitrogen", "cec", "clay", "sand")

re_weather <- c("days_hot_crit_gs", "days_cold_crit_gs",
                "sum_heat_pos_gs", "sum_cold_neg_gs",
                "total_precip_gs", "mean_wind_gs",
                "mean_soil_moisture_pct", "days_drought")

re_formula <- as.formula(paste(
  "yield_cpha ~",
  paste(c(re_weather, soil_vars, "log_sown", "war_east"), collapse = " + "),
  "+ (1 | region) + (1 | year)"
))

m_re <- lme4::lmer(re_formula, data = panel_base, REML = FALSE)
summary(m_re)

soil_coefs_re <- coef(summary(m_re)) |>
  as.data.frame() |>
  rownames_to_column("variable") |>
  filter(variable %in% soil_vars) |>
  rename(std_error = `Std. Error`, t_value = `t value`) |>
  mutate(
    p_approx    = 2 * pnorm(abs(t_value), lower.tail = FALSE),
    significant = case_when(
      p_approx < 0.01 ~ "***",
      p_approx < 0.05 ~ "**",
      p_approx < 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    Estimate  = round(Estimate,  3),
    std_error = round(std_error, 3)
  ) |>
  dplyr::select(variable, Estimate, std_error, t_value, p_approx, significant)

knitr::kable(soil_coefs_re, digits = 3,
  caption = "RE model — soil variable coefficients (approximate p-values via normal distribution)")

cat("\nJoint Wald test for all soil variables (RE model):\n")
soil_joint_re <- multcomp::glht(m_re, linfct = paste0(soil_vars, " = 0"))
summary(soil_joint_re)
```

------------------------------------------------------------------------

## 3.3.2 Mundlak-Chamberlain device

We compute within-region means of all time-varying weather regressors and add them to the RE model. The Wald test on these mean terms serves as a Hausman-type specification test.

```{r mundlak}
mundlak_vars <- re_weather

mundlak_means <- panel_base |>
  group_by(region) |>
  summarise(
    across(all_of(mundlak_vars),
           ~ mean(.x, na.rm = TRUE),
           .names = "{.col}_mean"),
    .groups = "drop"
  )

panel_mundlak <- panel_base |>
  left_join(mundlak_means, by = "region")

mundlak_mean_cols <- paste0(mundlak_vars, "_mean")

mundlak_formula <- as.formula(paste(
  "yield_cpha ~",
  paste(c(re_weather, soil_vars, mundlak_mean_cols,
          "log_sown", "war_east"), collapse = " + "),
  "+ (1 | region) + (1 | year)"
))

m_mundlak <- lmer(mundlak_formula, data = panel_mundlak, REML = FALSE)
summary(m_mundlak)

soil_coefs_mundlak <- coef(summary(m_mundlak)) |>
  as.data.frame() |>
  rownames_to_column("variable") |>
  filter(variable %in% soil_vars) |>
  rename(std_error = `Std. Error`, t_value = `t value`) |>
  mutate(
    p_approx    = 2 * pnorm(abs(t_value), lower.tail = FALSE),
    significant = case_when(
      p_approx < 0.01 ~ "***",
      p_approx < 0.05 ~ "**",
      p_approx < 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    Estimate  = round(Estimate,  3),
    std_error = round(std_error, 3)
  ) |>
  dplyr::select(variable, Estimate, std_error, t_value, p_approx, significant)

knitr::kable(soil_coefs_mundlak, digits = 3,
  caption = "Mundlak-Chamberlain model — soil variable coefficients")

cat("\nHausman-type test: joint significance of Mundlak mean terms\n")
cat("H₀: λ = 0 (RE orthogonality holds, plain RE is consistent)\n")
cat("H₁: λ ≠ 0 (RE is biased, Mundlak correction is needed)\n\n")
mundlak_joint <- glht(m_mundlak, linfct = paste0(mundlak_mean_cols, " = 0"))
summary(mundlak_joint)

cat("\nAIC comparison:\n")
cat("RE model AIC:      ", round(AIC(m_re), 2), "\n")
cat("Mundlak model AIC: ", round(AIC(m_mundlak), 2), "\n")
cat("Lower AIC is preferred.\n")
```

------------------------------------------------------------------------

## 3.3.3 Soil × weather interactions (OLS with region dummies)

We return to OLS with region dummies (the same framework as Parts 1 and 2) and add soil × weather interaction terms. These are identified under region dummies because the interaction varies over time. We test four theoretically motivated interactions:

-   `organic_matter × days_hot_crit_gs` — does richer soil buffer heat stress?
-   `organic_matter × days_drought` — does organic matter retain more plant-available water during drought?
-   `clay × days_hot_crit_gs` — does clay content (higher water retention) reduce heat damage?
-   `clay × days_drought` — does clay protect against drought stress?

```{r soil-interactions}
panel_interact <- panel_base |>
  mutate(
    om_x_hot       = organic_matter * days_hot_crit_gs,
    om_x_drought   = organic_matter * days_drought,
    clay_x_hot     = clay * days_hot_crit_gs,
    clay_x_drought = clay * days_drought
  )

interaction_cols <- c("om_x_hot", "om_x_drought", "clay_x_hot", "clay_x_drought")

fml_interact <- as.formula(paste(
  "yield_cpha ~",
  paste(c(re_weather, interaction_cols, all_controls), collapse = " + ")
))

m_interact <- lm(fml_interact, data = panel_interact)
summary(m_interact)

interact_coefs <- coef(summary(m_interact)) |>
  as.data.frame() |>
  rownames_to_column("variable") |>
  filter(variable %in% interaction_cols) |>
  rename(std_error = `Std. Error`, t_value = `t value`, p_value = `Pr(>|t|)`) |>
  mutate(
    label = case_when(
      variable == "om_x_hot"       ~ "Organic matter × Hot days (GS)",
      variable == "om_x_drought"   ~ "Organic matter × Drought days",
      variable == "clay_x_hot"     ~ "Clay × Hot days (GS)",
      variable == "clay_x_drought" ~ "Clay × Drought days"
    ),
    significant = case_when(
      p_value < 0.01 ~ "***",
      p_value < 0.05 ~ "**",
      p_value < 0.10 ~ "*",
      TRUE           ~ "ns"
    ),
    Estimate  = round(Estimate,  4),
    std_error = round(std_error, 4),
    p_value   = round(p_value,   4)
  ) |>
  dplyr::select(label, Estimate, std_error, t_value, p_value, significant)

knitr::kable(interact_coefs, digits = 4,
  caption = "Soil × weather interaction coefficients (OLS with region dummies)")

cat("\nJoint significance of all soil × weather interactions:\n")
linearHypothesis(m_interact,
  c("om_x_hot = 0", "om_x_drought = 0", "clay_x_hot = 0", "clay_x_drought = 0"))
```

------------------------------------------------------------------------

## 3.3.4 Individual and joint significance summary

```{r soil-significance-summary}
soil_compare <- soil_coefs_re |>
  dplyr::select(variable, Estimate, p_approx, significant) |>
  rename(est_RE = Estimate, p_RE = p_approx, sig_RE = significant) |>
  left_join(
    soil_coefs_mundlak |>
      dplyr::select(variable, Estimate, p_approx, significant) |>
      rename(est_Mundlak = Estimate, p_Mundlak = p_approx, sig_Mundlak = significant),
    by = "variable"
  )

knitr::kable(soil_compare, digits = 3,
  caption = "Soil variable coefficients: RE vs Mundlak-Chamberlain comparison")

cat("\n=== Joint significance summary ===\n")
cat("RE model — joint test p-value: see glht output above\n")
cat("Mundlak model — joint test p-value: see glht output above\n")
cat("Interaction model — joint test p-value: see linearHypothesis output above\n")
```

------------------------------------------------------------------------

## 3.3.5 Part 3 comparison table

```{r part3-comparison}
part3_table <- tibble(
  Model = c(
    "RE (soil direct effects)",
    "Mundlak-Chamberlain (soil + climate controls)",
    "OLS + region dummies + soil × weather interactions"
  ),
  N      = c(nobs(m_re), nobs(m_mundlak), nobs(m_interact)),
  Adj_R2 = c(NA_real_, NA_real_, summary(m_interact)$adj.r.squared),
  AIC    = c(AIC(m_re), AIC(m_mundlak), AIC(m_interact)),
  BIC    = c(BIC(m_re), BIC(m_mundlak), BIC(m_interact))
)

knitr::kable(part3_table, digits = 3,
  caption = "Part 3 model comparison: three approaches to soil variable identification")
```

*Note: Adj R² is not directly comparable between lme4 mixed models (RE, Mundlak) and OLS (interactions model). AIC and BIC are valid for comparison within each estimator class but should be interpreted with caution across classes.*

------------------------------------------------------------------------

## 3.3.6 Conclusions from soil analysis

**On the choice of estimator:**

The Hausman-type test on the Mundlak mean terms reveals whether the plain RE model is adequate. If the joint test on $\boldsymbol{\lambda}$ rejects the null, the RE orthogonality assumption fails — meaning that the long-run climate of each oblast is correlated with its unobserved random effect. In this case, the Mundlak model is preferred: it corrects for this bias by explicitly controlling for regional climate averages, and its soil coefficients should be interpreted as the effect of soil quality *conditional on the oblast's climate*.

**On direct soil effects (RE and Mundlak):**

The RE model provides a first approximation of which soil properties matter. Positive coefficients on `organic_matter` would confirm that higher soil carbon improves average yield — consistent with its role in water retention, nutrient supply, and microbial activity. The `clay` coefficient captures a trade-off: clay soils retain more moisture (positive in dry years) but also drain poorly and warm more slowly in spring (potentially negative overall). If the Mundlak correction substantially changes the magnitude or significance of soil coefficients relative to plain RE, this confirms that ignoring the climate-soil correlation would have led to misleading inferences.

**On soil × weather interactions (OLS model):**

The interaction coefficients answer the most actionable question for agribusiness: does soil quality change the *sensitivity* of yield to weather shocks? A positive and significant coefficient on `organic_matter × days_hot_crit_gs` means that each additional hot day causes less yield loss in oblasts with richer soils — direct empirical evidence for the buffering role of soil organic matter under heat stress. Similarly, a significant `organic_matter × days_drought` interaction would support the agronomic claim that organic matter improves drought resilience by enhancing water-holding capacity. If the joint test on all four interaction terms is significant, we can conclude that soil quality modulates weather risk — an important finding for climate adaptation strategy.

**Limitation:** With only 24 oblasts and static soil data, identification of soil effects relies entirely on cross-sectional variation. The estimates are sensitive to unobserved regional factors that co-vary with both soil quality and yield (e.g., management culture, land consolidation, irrigation access). The Mundlak correction partially addresses this but cannot eliminate all confounding. These results should be interpreted as indicative associations rather than causal estimates.

---

# Conclusions

## What the data showed

The exploratory analysis established several facts that directly shaped the modelling strategy. First, yield levels and trends differ substantially across oblasts — western regions consistently outperform eastern ones, and the patterns diverge enough that a single common trend cannot summarise all regions. This motivated including both a common time trend and, where warranted, region-specific trends. Second, the VIF analysis confirmed that solar radiation, ET₀, and dewpoint are largely redundant given the information already captured by temperature stress metrics and precipitation: removing these three variables reduced the maximum VIF from 70.25 to 6.91 with no loss of substantive content.

## What the models showed

**Part 1 — Stationarity and trend.** Raw yield series are non-stationary in most oblasts, but model residuals show no persistent unit-root behaviour, confirming that the regression specification adequately absorbs trends. Adding a common time trend improved adjusted R² from 0.744 to 0.764 and reduced AIC by 16 points. Region-specific trends added no further improvement and consumed too many degrees of freedom. The lagged loss rate was individually insignificant, suggesting that farmer behavioural responses to prior-year losses are not a primary driver of subsequent yield in this dataset.

**Part 2 — Weather window.** The growing season model (M_gs) outperformed the full-year model on all criteria: adjusted R² of 0.794 versus 0.764, AIC lower by 28 points. The two-split and three-split models performed similarly to M_gs but added complexity without meaningful gain. Within the growing season specification, the key significant weather variables were: days with extreme heat stress (positive, counterintuitive — likely a proxy for summer radiation), accumulated non-critical heat (negative — sustained warmth above the growth optimum reduces yield), total growing season precipitation (positive), and mean wind speed (negative — lodging risk). These results confirm Hypothesis 3: the growing season window dominates, supporting Hypothesis 2, and extreme-day counts carry predictive content beyond mean temperature, supporting Hypothesis 3.

**Part 3 — Soil variables.** Three complementary approaches were used to identify soil effects that are collinear with region dummies in the OLS framework. The plain RE model identified pH, clay, and sand as individually significant. After applying the Mundlak-Chamberlain correction — which controls for the correlation between long-run regional climate and soil quality — organic matter and sand emerged as the most robust soil effects, while clay lost significance. The Hausman-type test on the Mundlak mean terms was only marginally significant for two terms (accumulated cold and precipitation), suggesting the RE orthogonality assumption is not severely violated but the Mundlak model is still preferred on AIC grounds (1189 vs 1222). The soil × weather interaction model found no statistically significant interactions between organic matter or clay and heat or drought stress (joint F-test p = 0.61), failing to support the buffering hypothesis with this dataset.

## Answers to the research hypotheses

**H1 (meteorological significance):** Confirmed. Growing season temperature stress, precipitation, and wind are jointly and individually significant drivers of yield. The relationship is non-linear in the sense that both heat stress days and accumulated heat above the growth optimum matter, with opposing signs.

**H2 (seasonal relevancy):** Confirmed. The growing season model outperforms the full-year model. Winter weather variables, once separated into their own window, add little incremental explanatory power beyond the growing season.

**H3 (extreme day classification):** Confirmed. Days with extreme heat stress and accumulated non-critical heat are both significant, while raw mean temperature metrics have higher VIF and lower t-statistics. Daily data enabling extreme-event counting adds value over simple averages.

**H4 (redundancy):** Confirmed. Solar radiation, ET₀, and dewpoint can be dropped from the data collection protocol without loss. The mandatory set is: daily maximum and minimum temperature, daily precipitation, daily wind speed, and soil moisture.

**H5 (soil as moderator):** Partially confirmed. Soil organic matter and sand content have direct effects on average yield (identified through the Mundlak model), but do not significantly moderate weather sensitivity in the interaction model. Soil quality matters for baseline productivity but does not appear to buffer heat or drought shocks in this 9-year panel.

## Practical recommendation for agribusinesses

The minimum sufficient data collection protocol for yield forecasting at oblast level consists of: daily Tmax and Tmin (to construct extreme-day counts and accumulated heat/cold), daily precipitation totals, daily wind gusts, and periodic soil moisture readings at 0–7 cm depth. Solar radiation, ET₀, dewpoint, and cloudiness are redundant given this set and need not be collected. Soil properties from a single baseline survey provide useful information about regional yield potential but should not be expected to improve short-run weather-driven forecasts.
