### Collinearity diagnostics among regressors

The goal here is not to correlate variables with yield, but to check
whether the regressors correlate with each other in ways that would
make it impossible for OLS to separate their individual effects.
Two tools are used in combination because each catches what the other
misses:

- **Pearson r between regressor pairs** catches linear collinearity.
  It will flag ET0 and solar radiation (r = 0.89) but will not detect
  cases where two variables are related nonlinearly or where three
  variables jointly span the same dimension even though no single pair
  is highly correlated.

- **VIF (Variance Inflation Factor)** catches the full multicollinearity
  structure including joint redundancy. A variable can have moderate
  pairwise correlations with several others and still produce a very
  high VIF because it is collectively explained by the group.
  Rule of thumb: VIF > 5 warrants attention, VIF > 10 is severe.

```{r collinearity, fig.width = 10, fig.height = 9}
library(corrplot)
library(car)

# ── Regressor set: only variables entering the model ──────────
# Four temperature metrics + raw weather + harvest ratio
# Soil variables are time-invariant and will be absorbed by oblast
# fixed effects in the panel model, so they are excluded here.

regressors <- c(
  "days_hot_crit",
  "days_cold_crit",
  "sum_heat_noncrit",
  "sum_cold_noncrit",
  "total_precip",
  "total_et0",
  "total_radiation",
  "mean_wind",
  "mean_dewpoint",
  "harv_ratio"
)

# keep only columns that exist
regressors <- intersect(regressors, names(panel_full))

reg_df <- panel_full |>
  select(all_of(regressors)) |>
  drop_na()

cat("Observations used for collinearity diagnostics:", nrow(reg_df), "\n")

# ── Display labels ────────────────────────────────────────────
reg_labels <- c(
  days_hot_crit    = "Hot critical days",
  days_cold_crit   = "Cold critical days",
  sum_heat_noncrit = "Accum. heat",
  sum_cold_noncrit = "Accum. cold",
  total_precip     = "Precipitation",
  total_et0        = "ET0",
  total_radiation  = "Solar radiation",
  mean_wind        = "Wind gusts",
  mean_dewpoint    = "Dewpoint",
  harv_ratio       = "Harvest ratio"
)

reg_labels <- reg_labels[regressors]

# ── 1. Correlation matrix among regressors ────────────────────
cor_reg <- cor(reg_df, use = "pairwise.complete.obs")
rownames(cor_reg) <- colnames(cor_reg) <- reg_labels

par(bg = "white")
corrplot(
  cor_reg,
  method      = "color",
  type        = "upper",
  order       = "hclust",
  tl.col      = "grey15",
  tl.srt      = 40,
  tl.cex      = 0.85,
  cl.cex      = 0.75,
  addCoef.col = "grey10",
  number.cex  = 0.72,
  col         = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(200),
  title       = "Correlation matrix among regressors",
  mar         = c(0, 0, 2, 0)
)
```

```{r vif, fig.width = 7, fig.height = 5}
# ── 2. VIF for the full regressor set ─────────────────────────
# OLS without fixed effects — fixed effects are not needed here
# because VIF measures regressor redundancy, not causal identification.
# Adding a constant outcome variable (yield) is enough.

vif_formula <- as.formula(
  paste("yield_cpha ~", paste(regressors, collapse = " + "))
)

vif_data <- panel_full |>
  select(yield_cpha, all_of(regressors)) |>
  drop_na()

vif_model  <- lm(vif_formula, data = vif_data)
vif_values <- vif(vif_model)

vif_df <- tibble(
  variable = names(vif_values),
  VIF      = vif_values
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High — must address",
      VIF > 5  ~ "Moderate — monitor",
      TRUE     ~ "Acceptable"
    ),
    concern = factor(concern,
                     levels = c("Acceptable",
                                "Moderate — monitor",
                                "High — must address"))
  ) |>
  arrange(desc(VIF))

# table
knitr::kable(
  vif_df |> select(label, VIF, concern),
  digits  = 2,
  col.names = c("Variable", "VIF", "Assessment"),
  caption = "Variance Inflation Factors — full regressor set"
)
```

```{r vif-plot, fig.width = 8, fig.height = 5}
# VIF bar chart — easier to read than a table for the report
ggplot(vif_df,
       aes(x = VIF,
           y = fct_reorder(label, VIF),
           fill = concern)) +
  geom_col(width = 0.68, alpha = 0.88) +
  geom_text(aes(label = round(VIF, 1)),
            hjust = -0.15, size = 3, colour = "grey25") +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#d7191c", linewidth = 0.5) +
  annotate("text", x = 5.2,  y = 0.6, label = "VIF = 5",
           hjust = 0, size = 2.6, colour = "grey50") +
  annotate("text", x = 10.2, y = 0.6, label = "VIF = 10",
           hjust = 0, size = 2.6, colour = "#d7191c") +
  scale_fill_manual(values = c(
    "Acceptable"           = "#1a9641",
    "Moderate — monitor"   = "#fdae61",
    "High — must address"  = "#d7191c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title   = "Variance Inflation Factors — full regressor set",
    x       = "VIF", y = NULL, fill = NULL,
    caption = "Dashed grey line = VIF 5 threshold. Dashed red line = VIF 10 threshold."
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    legend.position    = "top"
  )
```

```{r vif-reduced, fig.width = 8, fig.height = 5}
# ── 3. VIF after dropping the most collinear variables ─────────
# If ET0 and solar radiation both produce high VIF, drop solar radiation
# (absorbed by ET0) and recheck. This iterative approach shows clearly
# which variable is causing the inflation.

regressors_reduced <- regressors[!regressors %in% c("total_radiation",
                                                      "mean_dewpoint")]

vif_formula2 <- as.formula(
  paste("yield_cpha ~", paste(regressors_reduced, collapse = " + "))
)

vif_model2  <- lm(vif_formula2,
                  data = panel_full |>
                    select(yield_cpha, all_of(regressors_reduced)) |>
                    drop_na())
vif_values2 <- vif(vif_model2)

vif_df2 <- tibble(
  variable = names(vif_values2),
  VIF      = vif_values2
) |>
  mutate(
    label   = coalesce(reg_labels[variable], variable),
    concern = case_when(
      VIF > 10 ~ "High — must address",
      VIF > 5  ~ "Moderate — monitor",
      TRUE     ~ "Acceptable"
    ),
    concern = factor(concern,
                     levels = c("Acceptable",
                                "Moderate — monitor",
                                "High — must address"))
  ) |>
  arrange(desc(VIF))

ggplot(vif_df2,
       aes(x = VIF,
           y = fct_reorder(label, VIF),
           fill = concern)) +
  geom_col(width = 0.68, alpha = 0.88) +
  geom_text(aes(label = round(VIF, 1)),
            hjust = -0.15, size = 3, colour = "grey25") +
  geom_vline(xintercept = 5,  linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 10, linetype = "dashed",
             colour = "#d7191c", linewidth = 0.5) +
  annotate("text", x = 5.2,  y = 0.6, label = "VIF = 5",
           hjust = 0, size = 2.6, colour = "grey50") +
  annotate("text", x = 10.2, y = 0.6, label = "VIF = 10",
           hjust = 0, size = 2.6, colour = "#d7191c") +
  scale_fill_manual(values = c(
    "Acceptable"           = "#1a9641",
    "Moderate — monitor"   = "#fdae61",
    "High — must address"  = "#d7191c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title   = "VIF after removing solar radiation and dewpoint",
    subtitle = "Shows whether removing the most collinear variables resolves inflation in the remaining set.",
    x       = "VIF", y = NULL, fill = NULL,
    caption = "Dashed grey line = VIF 5. Dashed red line = VIF 10."
  ) +
  theme_small_multiple(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    axis.line.x        = element_blank(),
    legend.position    = "top"
  )
```
