---
title: "Weather Variables – Exploratory Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 7, dpi = 150)
library(tidyverse)
library(lubridate)
library(patchwork)
```

## 3.1 Why these temperature metrics?

Winter wheat passes through several development stages — vernalisation, tillering, stem extension, heading, and grain filling — each with a different thermal optimum. A single summary statistic such as the annual mean temperature loses most of the agronomically relevant variation. The decomposition used here is motivated by the following reasoning:

**Base temperature (5 °C).** Below 5 °C wheat development effectively halts. We therefore express all thermal signals as deviations from this threshold: $t_{\text{dev}} = \frac{T_{\max} + T_{\min}}{2} - 5$.

**Why both maximum and minimum?**\
Daily maximum temperature governs daytime photosynthesis, grain filling rate, and heat stress at flowering. Daily minimum drives overnight respiration costs and frost damage risk. Using their arithmetic mean captures the full diurnal thermal load rather than relying on a single extreme or an imprecise mean that could be derived from very different combinations of highs and lows.

**Critical hot days** ($t_{\text{dev}} > 20$, i.e. mean day ≥ 25 °C).\
Above this threshold photosynthesis is impaired, pollen viability drops sharply, and grain-filling is shortened. Each such day can cause irreversible yield loss. We count them — not sum them — because the *number of stress episodes* matters more than their cumulative magnitude once the damage threshold is crossed.

**Critical cold days** ($t_{\text{dev}} < -15$, i.e. mean day ≤ −10 °C).\
Severe frost during winter dormancy or early spring can kill the crop outright. Again, *count* of such days is the relevant signal.

**Accumulated non-critical heat** (sum of $t_{\text{dev}}$ for $0 < t_{\text{dev}} \leq 20$).\
This is the growing-degree-day budget — the thermal energy available for vegetative growth and grain development. A larger accumulated sum generally indicates a more productive season, provided heat stress days remain low.

**Accumulated non-critical cold** (sum of $t_{\text{dev}}$ for $-15 \leq t_{\text{dev}} < 0$).\
Mild sub-zero temperatures during vernalisation are *required* by winter wheat. Insufficient chilling leads to incomplete vernalisation and reduced tillering. This variable captures whether winters are cold enough — a deficit (less negative sum) may be as harmful as an excess.

------------------------------------------------------------------------

```{r load-weather}
# ── Load and aggregate ─────────────────────────────────────
weather <- read_csv("ukraine_daily_weather.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date), year = year(date)) |>
  filter(year >= 2013, year <= 2024)

BASE_TEMP <- 5

weather_annual <- weather |>
  mutate(
    t_mean         = (temperature_2m_max + temperature_2m_min) / 2,
    t_dev          = t_mean - BASE_TEMP,
    days_hot_crit  = t_dev > 20,
    days_cold_crit = t_dev < -15,
    heat_noncrit   = if_else(t_dev > 0  & t_dev <= 20,  t_dev, 0),
    cold_noncrit   = if_else(t_dev >= -15 & t_dev < 0,  t_dev, 0)
  ) |>
  group_by(region, year) |>
  summarise(
    days_hot_crit    = sum(days_hot_crit,  na.rm = TRUE),
    days_cold_crit   = sum(days_cold_crit, na.rm = TRUE),
    sum_heat_noncrit = sum(heat_noncrit,   na.rm = TRUE),
    sum_cold_noncrit = sum(cold_noncrit,   na.rm = TRUE),
    mean_tmax        = mean(temperature_2m_max,          na.rm = TRUE),
    mean_tmin        = mean(temperature_2m_min,          na.rm = TRUE),
    mean_temp        = mean(t_mean,                      na.rm = TRUE),
    total_precip     = sum(precipitation_sum,            na.rm = TRUE),
    mean_et0         = mean(et0_fao_evapotranspiration,  na.rm = TRUE),
    total_et0        = sum(et0_fao_evapotranspiration,   na.rm = TRUE),
    mean_wind        = mean(wind_gusts_10m_max,          na.rm = TRUE),
    .groups          = "drop"
  )

# Order regions W→E by mean temperature (warmer = further east/south)
region_order <- weather_annual |>
  group_by(region) |>
  summarise(mt = mean(mean_temp)) |>
  arrange(mt) |>
  pull(region)

weather_annual <- weather_annual |>
  mutate(region = factor(region, levels = region_order),
         year   = factor(year))
```

```{r theme}
# ── Shared theme ────────────────────────────────────────────
theme_heatmap <- function(base_size = 9) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid        = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      axis.text.x       = element_text(size = 7.5, angle = 45, hjust = 1,
                                       colour = "grey35"),
      axis.text.y       = element_text(size = 7,   colour = "grey30"),
      axis.title        = element_blank(),
      axis.ticks        = element_blank(),
      strip.text        = element_text(face = "bold", size = 8.5,
                                       colour = "grey15"),
      strip.background  = element_blank(),
      plot.title        = element_text(face = "bold", size = 10,
                                       colour = "grey10",
                                       margin = margin(b = 2)),
      plot.subtitle     = element_text(size = 7.5, colour = "grey45",
                                       margin = margin(b = 5)),
      legend.text       = element_text(size = 7,   colour = "grey35"),
      legend.title      = element_blank(),
      legend.key.width  = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin       = margin(6, 8, 6, 6)
    )
}

# ── Helper: one heatmap tile ────────────────────────────────
make_tile <- function(df, metric, palette, direction = 1,
                      title, subtitle) {
  ggplot(df, aes(x = year, y = region, fill = .data[[metric]])) +
    geom_tile(colour = "white", linewidth = 0.35) +
    scale_fill_distiller(palette   = palette,
                         direction = direction,
                         guide     = guide_colorbar(
                           barwidth  = 0.45,
                           barheight = 7,
                           ticks     = FALSE)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(title = title, subtitle = subtitle) +
    theme_heatmap()
}
```

------------------------------------------------------------------------

### 3.2 Temperature stress metrics

The four panels below show each metric as a **region × year heatmap**. This format is chosen deliberately: it reveals both the *spatial* gradient (warmer eastern/southern oblasts appear at the top) and the *temporal* trend (recent warming is visible as a rightward intensification of heat-stress tiles). Each cell aggregates all daily observations for that oblast-year combination, so the 70 000+ daily rows are reduced to the \~288 cells that actually carry the agronomic signal.

```{r temp-stress, fig.height=8}
p1 <- make_tile(weather_annual, "days_hot_crit", "YlOrRd", 1,
  "Days with extreme heat stress",
  "t_dev > 20 °C  (mean daily temp > 25 °C) — count of days per year")

p2 <- make_tile(weather_annual, "days_cold_crit", "Blues", -1,
  "Days with extreme cold stress",
  "t_dev < −15 °C  (mean daily temp < −10 °C) — count of days per year")

p3 <- make_tile(weather_annual, "sum_heat_noncrit", "YlOrBr", 1,
  "Accumulated non-critical heat  (growing degree-days)",
  "Sum of t_dev for days where 0 < t_dev ≤ 20 °C — thermal budget for growth")

p4 <- make_tile(weather_annual, "sum_cold_noncrit", "PuBu", -1,
  "Accumulated non-critical cold  (vernalisation budget)",
  "Sum of t_dev for days where −15 ≤ t_dev < 0 °C — chilling requirement")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Temperature stress metrics by region and year",
    subtitle = paste0("Derived from daily Tmax / Tmin. Base temperature = ",
                      BASE_TEMP, " °C. Regions ordered cool → warm (W/N to E/S)."),
    caption  = "Source: ukraine_daily_weather.csv",
    theme    = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.subtitle   = element_text(size = 9, colour = "grey45"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

**Key observations:**

-   Heat stress days are concentrated in eastern and southern oblasts and have increased noticeably after 2020, consistent with recent warming trends in Ukraine.
-   Cold stress days are rare across most of the study period, suggesting that extreme frost winters are uncommon — but note that even a single such day at the wrong phenological stage can be catastrophic.
-   The growing degree-day budget (non-critical heat) is relatively stable over time and shows a clear S–N gradient, confirming that the East/South regions are thermally advantaged for wheat production.
-   The vernalisation budget (non-critical cold) shows a declining trend in more recent years in southern regions, which could compromise vernalisation fulfilment and ultimately reduce yield potential.

------------------------------------------------------------------------

### 3.3 Raw temperature levels (Tmax and Tmin)

While the derived stress metrics above are the primary explanatory variables, it is useful to confirm that the spatial and temporal gradients in raw temperatures are as expected. The panels below show mean annual Tmax and Tmin per region and year.

```{r raw-temp}
p_tmax <- make_tile(weather_annual, "mean_tmax", "RdYlBu", -1,
  "Mean daily Tmax  (°C)",
  "Annual average of daily maximum temperature")

p_tmin <- make_tile(weather_annual, "mean_tmin", "RdYlBu", -1,
  "Mean daily Tmin  (°C)",
  "Annual average of daily minimum temperature")

p_tmean <- make_tile(weather_annual, "mean_temp", "RdYlBu", -1,
  "Mean daily temperature  (°C)",
  "Annual average of (Tmax + Tmin) / 2")

(p_tmax | p_tmin | p_tmean) +
  plot_annotation(
    title   = "Raw temperature levels by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

The warm-colour cells in the top rows (eastern/southern oblasts) and the rightward warming signal after 2019–2020 corroborate the patterns seen in the derived stress metrics. The Tmax–Tmin spread (diurnal range) is wider in continental eastern oblasts than in the more maritime west — relevant because a large diurnal range can mask both heat and cold stress when only a mean temperature is reported.

------------------------------------------------------------------------

### 3.4 Precipitation

Total annual precipitation is an important co-variate: heat stress is substantially amplified under drought conditions. The map below shows that the spatial gradient is roughly the inverse of the temperature gradient — wetter in the west and north-west, drier in the east and south — which compounds the heat stress disadvantage already visible in eastern oblasts.

```{r precip}
make_tile(weather_annual, "total_precip", "BrBG", 1,
  "Total annual precipitation  (mm)",
  "Sum of daily precipitation. Green = wetter, brown = drier.") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

Notable features:

-   A sharp dip in precipitation is visible in several eastern regions around 2020–2022, coinciding with the hottest years in the heat-stress panels — a compounding stress signal.
-   Western oblasts show comparatively stable and higher precipitation, consistent with Atlantic moisture influence.

------------------------------------------------------------------------

### 3.5 Evapotranspiration (ET₀)

Reference evapotranspiration (FAO-56 Penman–Monteith method) integrates temperature, radiation, wind, and humidity into a single measure of atmospheric water demand. High ET₀ combined with low precipitation defines agronomic drought — the conditions most damaging to wheat yield during stem extension and grain filling.

```{r et0}
p_et0_mean <- make_tile(weather_annual, "mean_et0", "YlOrRd", 1,
  "Mean daily ET₀  (mm/day)",
  "Annual average reference evapotranspiration")

p_et0_total <- make_tile(weather_annual, "total_et0", "YlOrRd", 1,
  "Total annual ET₀  (mm/year)",
  "Sum of daily reference evapotranspiration")

(p_et0_mean | p_et0_total) +
  plot_annotation(
    title   = "Reference evapotranspiration (ET₀) by region and year",
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.title      = element_text(face = "bold", size = 13, colour = "grey10"),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0),
      plot.background = element_rect(fill = "white", colour = NA)
    )
  )
```

ET₀ mirrors the temperature gradient — eastern regions show consistently higher atmospheric demand. The post-2019 intensification is clearly visible, aligning with the heat-stress trends. ET₀ will likely appear as a strong predictor in the regression models, potentially capturing drought stress effects that temperature alone cannot fully represent.

------------------------------------------------------------------------

### 3.6 Wind gusts

Maximum wind gusts can cause lodging (stem breakage) in wheat, which reduces harvestable yield independently of thermal or water stress. Wind also accelerates evapotranspiration and can amplify topsoil moisture loss during dry periods.

```{r wind}
make_tile(weather_annual, "mean_wind", "PuRd", 1,
  "Mean annual maximum wind gusts  (m/s)",
  "Annual average of daily maximum wind gust at 10 m height") +
  plot_annotation(
    caption = "Source: ukraine_daily_weather.csv",
    theme   = theme(
      plot.background = element_rect(fill = "white", colour = NA),
      plot.caption    = element_text(size = 7.5, colour = "grey55", hjust = 0)
    )
  )
```

Wind patterns show less clear spatial structure than temperature or precipitation. However, steppes oblasts (Zaporizhzhia, Kherson, Mykolaiv) tend to record higher gusts, consistent with their open landscape exposure. Wind will be included as a control variable in the regression models rather than a primary explanatory variable.

------------------------------------------------------------------------

### 3.7 Summary table of weather variable means (2013–2024)

```{r summary-table}
weather_annual |>
  group_by(region) |>
  summarise(
    `Hot days`     = round(mean(days_hot_crit),    1),
    `Cold days`    = round(mean(days_cold_crit),   1),
    `Heat sum`     = round(mean(sum_heat_noncrit), 0),
    `Cold sum`     = round(mean(sum_cold_noncrit), 0),
    `Precip (mm)`  = round(mean(total_precip),     0),
    `ET₀ (mm/yr)`  = round(mean(total_et0),        0),
    `Wind (m/s)`   = round(mean(mean_wind),        1)
  ) |>
  arrange(desc(`Hot days`)) |>
  knitr::kable(caption = "Mean annual weather metrics by oblast (2013–2024, ordered by heat stress days)")
```
